---
title: "13.4 - WGBS DNA methylation across three populations"
output:
  html_document:
    toc: true
    toc_float: true
date: "2026-10-08"
---

Population differences in CpG methylation in *Ostrea chilensis*, from whole-genome
bisulfite sequencing of oysters from Pumalin, Quilhua, and Rio Pudeto.

**Sample set.** 15 libraries were delivered (5 per site), but several FASTQ pairs
share reads with others (`output/13-wgbs/raw_read_overlap.tsv`; see README Known
issues). After removing duplicates, `data/13-wgbs-samples.tsv` holds 9 independent
oysters: Pum1, Qui1-4, Rio1-4. Only populations with at least `min_pop_n` oysters
are tested, so Pumalin (n = 1) is left out of this notebook until its libraries are
re-delivered; the contrasts below are built from whichever populations remain.

**Upstream:** `code/13-wgbs-submit.sh` (13.1 genome prep → 13.2 trim/align/extract
→ 13.3 QC summary) must be finished. This notebook reads the strand-merged CpG
files `output/13-wgbs/methylation/*.CpG_report.merged_CpG_evidence.cov.gz`.

**Reference:** `merged_out.fasta` + `GN.gene.gff3`, the same as the current RNA-seq
pipeline (`09`), so differentially methylated genes and DE genes share IDs.

**Where to run:** on a compute node inside the lab R container, e.g.

```
salloc -A srlab -p cpu-g2-mem2x -c 16 --mem=150G -t 12:00:00
apptainer exec /gscratch/srlab/containers/srlab-R4.4-bioinformatics-container-c3d3116.sif R
```

Paths are relative to `code/`, like the other notebooks.

**Depth caveat.** At ~23-38 M read pairs per sample, raw coverage of the 1.18 Gb
genome is only ~6-9x before mapping and deduplication. Single-CpG tests will
therefore be underpowered and cover a fraction of sites. The analysis is layered
so that the region (tile) and gene-body levels, which pool CpGs, carry most of the
inference; treat single-CpG results as supporting.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r packages}
library(methylKit)
library(GenomicRanges)
library(rtracklayer)
library(limma)
library(pheatmap)
library(tidyverse)
# tidyverse is loaded last, so it masks methylKit::unite (tidyr), and the range
# versions of reduce (purrr) and setdiff (dplyr/lubridate); those calls are namespaced.
```

```{r parameters}
out_dir   <- "../output/13-wgbs/13.4-methylation"
meth_dir  <- "../output/13-wgbs/methylation"
genome_dir <- "../data/13-wgbs-genome"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

min_cov        <- 5      # per-CpG minimum (strand-merged); 10 is typical but costs most sites here
max_cov_pctl   <- 99.9   # drop PCR/repeat pile-ups
min_pop_n      <- 3L     # populations with fewer oysters are not tested
min_per_group  <- 3L     # CpG must pass in >= 3 oysters (of 4) in every population
tile_size      <- 1000   # bp, non-overlapping windows for DMRs
tile_min_cpgs  <- 5      # CpGs required per tile per sample
min_gene_cpgs  <- 5      # CpGs required per gene for gene-body methylation
diff_cutoff    <- 20     # % methylation difference for DML/DMR calls
q_cutoff       <- 0.01
cores          <- 8
```

# Samples

Groups come from the explicit table `data/13-wgbs-samples.tsv`, never from file
order, so a re-sorted directory can't silently swap populations.

```{r samples}
samples_all <- read_tsv("../data/13-wgbs-samples.tsv", show_col_types = FALSE)
pop_n <- count(samples_all, population)
untested <- filter(pop_n, n < min_pop_n)
if (nrow(untested) > 0)
  message("Not tested (fewer than ", min_pop_n, " oysters): ",
          paste0(untested$population, " (n = ", untested$n, ")", collapse = ", "))

samples <- samples_all %>%
  filter(!population %in% untested$population) %>%
  mutate(population = factor(population, levels = intersect(c("Pum", "Qui", "Rio"), population)),
         treatment  = as.integer(population) - 1L,          # 0, 1, ... for methylKit
         file       = file.path(meth_dir,
                                paste0(sample, ".CpG_report.merged_CpG_evidence.cov.gz")))

stopifnot(all(file.exists(samples$file)), nlevels(samples$population) >= 2)
samples %>% select(sample, population, site)
```

# Alignment QC

From 13.3. CHH methylation is the conversion check: invertebrate non-CpG
methylation is near zero, so CHH % ≈ 100 − conversion efficiency. Flag any sample
above ~1%.

```{r align-qc}
aln <- read_tsv("../output/13-wgbs/alignment_summary.tsv", show_col_types = FALSE) %>%
  inner_join(select(samples_all, sample, population), by = "sample")   # clean samples only
aln

aln %>%
  pivot_longer(c(mapping_pct, dup_pct, CpG_meth_pct, CHH_meth_pct)) %>%
  ggplot(aes(sample, value, fill = population)) +
  geom_col() +
  facet_wrap(~name, scales = "free_y") +
  theme_bw() + theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

# Read CpG calls

`bismarkCoverage` format: chr, start, end, % methylation, count methylated,
count unmethylated (1-based). Strands are already merged, so no `destrand` later.

```{r read}
meth_raw <- methRead(
  location  = as.list(samples$file),
  sample.id = as.list(samples$sample),
  treatment = samples$treatment,
  assembly  = "Och_merged_out",
  pipeline  = "bismarkCoverage",
  context   = "CpG",
  mincov    = 1
)
```

```{r coverage-stats, fig.height=8}
# Coverage distribution per sample, before filtering.
cov_summary <- map_dfr(seq_along(meth_raw), function(i) {
  d <- getData(meth_raw[[i]])
  tibble(sample   = samples$sample[i],
         cpgs_1x  = nrow(d),
         cpgs_5x  = sum(d$coverage >= 5),
         cpgs_10x = sum(d$coverage >= 10),
         median_cov = median(d$coverage),
         # coverage-weighted global CpG methylation
         global_meth_pct = 100 * sum(d$numCs) / sum(d$coverage))
})
write_tsv(cov_summary, file.path(out_dir, "cpg_coverage_summary.tsv"))
cov_summary

ggplot(left_join(cov_summary, samples, by = "sample"),
       aes(population, global_meth_pct, colour = population)) +
  geom_jitter(width = 0.1, size = 3) +
  labs(y = "Global CpG methylation (%)", title = "Global methylation by population") +
  theme_bw()
```

# Filter, normalize, unite

```{r filter-unite}
meth_filt <- filterByCoverage(meth_raw, lo.count = min_cov, hi.perc = max_cov_pctl)
meth_norm <- normalizeCoverage(meth_filt, method = "median")

meth_unite <- methylKit::unite(meth_norm, destrand = FALSE,
                               min.per.group = min_per_group, mc.cores = cores)
message(nrow(meth_unite), " CpGs pass in >= ", min_per_group, " oysters per population")

pm <- percMethylation(meth_unite)          # CpG x sample, NAs where a sample was filtered
```

## Distribution

Invertebrate methylomes are typically mosaic/bimodal: most CpGs near 0% or near 100%,
with methylation concentrated in gene bodies.

```{r distribution}
as_tibble(pm) %>%
  pivot_longer(everything(), names_to = "sample", values_to = "pct") %>%
  filter(!is.na(pct)) %>%
  left_join(samples, by = "sample") %>%
  ggplot(aes(pct, colour = population, group = sample)) +
  geom_density() +
  labs(x = "% methylation per CpG") + theme_bw()
```

## Sample structure

PCA on CpGs covered in every tested oyster, using the most variable sites.

```{r pca}
pm_complete <- pm[complete.cases(pm), ]
top <- head(order(apply(pm_complete, 1, var), decreasing = TRUE), 50000)
pca <- prcomp(t(pm_complete[top, ]))
ve  <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

as_tibble(pca$x[, 1:3], rownames = "sample") %>%
  left_join(samples, by = "sample") %>%
  ggplot(aes(PC1, PC2, colour = population, label = sample)) +
  geom_point(size = 3) + geom_text(vjust = -0.8, size = 3, show.legend = FALSE) +
  labs(x = paste0("PC1 (", ve[1], "%)"), y = paste0("PC2 (", ve[2], "%)"),
       subtitle = paste(nrow(pm_complete), "complete CpGs; top", length(top), "by variance")) +
  theme_bw()

pheatmap(cor(pm_complete[top, ], method = "spearman"),
         annotation_col = data.frame(population = samples$population,
                                     row.names = samples$sample),
         main = "Spearman correlation, top variable CpGs")
```

# Genomic features

Annotation is built directly from `GN.gene.gff3`. Each CpG is assigned to one feature
by priority exon > intron > promoter (2 kb upstream of a gene) > intergenic.

```{r features}
seqlens <- read_tsv(file.path(genome_dir, "genome.chrom.sizes"),
                    col_names = c("chr", "len"), show_col_types = FALSE)
si <- Seqinfo(seqlens$chr, seqlens$len)

gff   <- import(file.path(genome_dir, "GN.gene.gff3"))
seqlevels(gff, pruning.mode = "coarse") <- seqlevels(si)   # align to FASTA order
seqinfo(gff) <- si                                            # lengths needed for trim()/gaps()

genes <- gff[gff$type == "gene"]
names(genes) <- genes$ID
exons     <- IRanges::reduce(gff[gff$type == "exon"], ignore.strand = TRUE)
introns   <- BiocGenerics::setdiff(IRanges::reduce(genes, ignore.strand = TRUE), exons)
upstream  <- trim(promoters(genes, upstream = 2000, downstream = 0))
promoters <- BiocGenerics::setdiff(IRanges::reduce(upstream, ignore.strand = TRUE),
                                   IRanges::reduce(genes, ignore.strand = TRUE))
intergenic <- gaps(IRanges::reduce(c(granges(genes), granges(promoters)), ignore.strand = TRUE))
intergenic <- intergenic[strand(intergenic) == "*"]

feature_sets <- list(exon = exons, intron = introns,
                     promoter = promoters, intergenic = intergenic)

annotate_feature <- function(gr) {
  out <- rep(NA_character_, length(gr))
  for (f in names(feature_sets)) {
    hit <- is.na(out) & overlapsAny(gr, feature_sets[[f]], ignore.strand = TRUE)
    out[hit] <- f
  }
  factor(out, levels = names(feature_sets))
}

cpg_gr <- as(meth_unite, "GRanges")
cpg_gr$feature <- annotate_feature(cpg_gr)
cpg_gr$mean_meth <- rowMeans(pm, na.rm = TRUE)

as_tibble(mcols(cpg_gr)) %>%
  ggplot(aes(feature, mean_meth)) +
  geom_violin(fill = "grey80") +
  stat_summary(fun = mean, geom = "point", colour = "red") +
  labs(y = "Mean % methylation across samples") + theme_bw()

table(cpg_gr$feature)
```

# Differential methylation

Every pairwise contrast among the tested populations, named as in the RNA-seq DE
(`Pum_vs_Qui`, `Pum_vs_Rio`, `Qui_vs_Rio`). Positive `meth.diff` = higher in the
second population.

```{r contrasts}
pairs <- combn(levels(samples$population), 2, simplify = FALSE)
contrasts <- set_names(pairs, map_chr(pairs, paste, collapse = "_vs_"))
names(contrasts)

# methylKit's binomial test with overdispersion correction, on a 2-population subset.
run_pair <- function(obj, groups) {
  keep <- samples %>% filter(population %in% groups)
  sub  <- reorganize(obj,
                     sample.ids = keep$sample,
                     treatment  = as.integer(keep$population == groups[2]))
  calculateDiffMeth(sub, overdispersion = "MN", test = "Chisq", mc.cores = cores)
}

write_hits <- function(diff, name, level) {
  hits <- getMethylDiff(diff, difference = diff_cutoff, qvalue = q_cutoff, type = "all")
  gr   <- as(hits, "GRanges")
  gr$feature <- annotate_feature(gr)
  write_tsv(as_tibble(as.data.frame(gr)),
            file.path(out_dir, paste0(level, "_", name, ".tsv")))
  export.bed(gr, file.path(out_dir, paste0(level, "_", name, ".bed")))
  gr
}
```

## Single CpGs (DMLs)

```{r dml}
dml <- imap(contrasts, function(g, name) {
  d <- run_pair(meth_unite, g)
  write_hits(d, name, "DML")
})
map_int(dml, length)
```

## 1 kb tiles (DMRs)

Tiling pools reads across neighbouring CpGs, which is the better-powered test at
this depth.

```{r dmr}
tiles <- tileMethylCounts(meth_norm, win.size = tile_size, step.size = tile_size,
                          cov.bases = tile_min_cpgs, mc.cores = cores)
tiles_unite <- methylKit::unite(tiles, destrand = FALSE, min.per.group = min_per_group,
                                mc.cores = cores)
message(nrow(tiles_unite), " tiles tested")

dmr <- imap(contrasts, function(g, name) {
  d <- run_pair(tiles_unite, g)
  write_hits(d, name, "DMR")
})

dmr_summary <- imap_dfr(dmr, ~ tibble(contrast = .y,
                                      n_dmr = length(.x),
                                      hyper = sum(.x$meth.diff > 0),
                                      hypo  = sum(.x$meth.diff < 0))) %>%
  left_join(imap_dfr(dmr, ~ as_tibble(table(feature = .x$feature)) %>%
                       mutate(contrast = .y)) %>%
              pivot_wider(names_from = feature, values_from = n),
            by = "contrast")
write_tsv(dmr_summary, file.path(out_dir, "DMR_summary.tsv"))
dmr_summary
```

# Gene-body methylation

Per gene and oyster: methylated / total calls summed over all CpGs in the gene
(a coverage-weighted mean). Genes need at least `min_gene_cpgs` CpGs with data
in every sample. Tested with limma on M-values (logit), the same framework and
contrasts as the RNA-seq DE in `09`, so results can be joined on gene ID.

```{r gene-body}
cov_mat <- as.matrix(getData(meth_unite)[, meth_unite@coverage.index])
cs_mat  <- as.matrix(getData(meth_unite)[, meth_unite@numCs.index])
colnames(cov_mat) <- colnames(cs_mat) <- meth_unite@sample.ids
cov_mat[is.na(cov_mat)] <- 0
cs_mat[is.na(cs_mat)]   <- 0

ov <- findOverlaps(cpg_gr, genes, ignore.strand = TRUE)
gene_id <- names(genes)[subjectHits(ov)]

sum_by_gene <- function(m) rowsum(m[queryHits(ov), , drop = FALSE], gene_id)
gene_cov <- sum_by_gene(cov_mat)
gene_cs  <- sum_by_gene(cs_mat)
gene_ncpg <- sum_by_gene((cov_mat > 0) * 1)

keep <- apply(gene_ncpg, 1, min) >= min_gene_cpgs
gene_frac <- (gene_cs[keep, ] + 0.5) / (gene_cov[keep, ] + 1)   # offset avoids logit(0/1)
gene_M    <- log2(gene_frac / (1 - gene_frac))
message(sum(keep), " genes with >= ", min_gene_cpgs, " covered CpGs in every oyster")

write_tsv(as_tibble(100 * gene_frac, rownames = "gene_id"),
          file.path(out_dir, "gene_body_methylation_pct.tsv"))

# Gene-body methylation is typically bimodal in bivalves (methylated vs unmethylated genes).
tibble(mean_pct = rowMeans(100 * gene_frac)) %>%
  ggplot(aes(mean_pct)) + geom_histogram(bins = 50) +
  labs(x = "Mean gene-body methylation (%)", y = "Genes") + theme_bw()
```

```{r gene-body-limma}
group  <- samples$population[match(colnames(gene_M), samples$sample)]
design <- model.matrix(~ 0 + group); colnames(design) <- levels(group)
cm <- makeContrasts(contrasts = map_chr(contrasts, ~ paste(.x[2], "-", .x[1])),
                    levels = design)
colnames(cm) <- names(contrasts)

fit <- eBayes(contrasts.fit(lmFit(gene_M, design), cm), trend = TRUE)

# Omnibus: any difference among the tested populations (with two populations
# this is just their single contrast).
any_pop <- topTable(fit, coef = if (ncol(cm) == 1) 1 else NULL, number = Inf,
                    sort.by = if (ncol(cm) == 1) "P" else "F") %>%
  as_tibble(rownames = "gene_id")
write_tsv(any_pop, file.path(out_dir, "DMG_any_population_F.tsv"))

dmg <- map(colnames(cm), function(cn) {
  tt <- topTable(fit, coef = cn, number = Inf) %>% as_tibble(rownames = "gene_id")
  # logFC here is a difference in M-values; also report it in % methylation.
  tt <- tt %>% mutate(delta_pct = 100 * (rowMeans(gene_frac[gene_id, group == sub(".*_vs_", "", cn)]) -
                                         rowMeans(gene_frac[gene_id, group == sub("_vs_.*", "", cn)])))
  write_tsv(tt, file.path(out_dir, paste0("DMG_", cn, ".tsv")))
  filter(tt, adj.P.Val < 0.05, abs(delta_pct) >= 10)
}) %>% set_names(colnames(cm))

map_int(dmg, nrow)

sig <- any_pop %>% filter(adj.P.Val < 0.05) %>% pull(gene_id)
if (length(sig) >= 2) {
  pheatmap(100 * gene_frac[head(sig, 100), ],
           scale = "row", show_rownames = length(sig) <= 50,
           annotation_col = data.frame(population = group, row.names = colnames(gene_M)),
           main = "Top differentially methylated genes (F-test)")
}
```

# Next steps

- **Expression link:** join `DMG_*.tsv` to the `09` limma tables on `gene_id`
  (GN gene IDs; the `09` StringTie gene IDs map to these via the reference-guided
  merge) and test whether gene-body methylation level or change predicts expression.
  Note the RNA-seq oysters are different individuals from the WGBS oysters, so this
  is a population-level, not individual-level, comparison.
- **Genetic confounding:** C→T SNPs read as unmethylated CpGs. Population-specific
  SNPs can create apparent DMLs. If many DMLs are near 0 vs 100%, call SNPs from
  the bisulfite BAMs (e.g. BS-SNPer / Biscuit) and drop CpGs overlapping C/T or G/A variants.
- **Functional enrichment:** GO enrichment of DMGs against the background of all
  tested genes (`any_pop$gene_id`), using `SPannot-GO.tsv` via the `03` join.

```{r session}
sessionInfo()
```
