---
title: "13.6 - Genome-wide DNA methylation landscape"
output:
  html_document:
    toc: true
    toc_float: true
date: "2026-10-08"
params:
  ref: "merged"   # or "hapa": the WGBS_REF=hapa alignment, with 13.11 lifted genes
---

General characterization of CpG methylation in *Ostrea chilensis*: how much of the
genome is methylated, how methylation is distributed across genomic features and
along chromosomes, what gene-body methylation looks like, and whether gene
methylation tracks gene function. Population differences are in `13.4`; this
notebook describes the shared methylome, pooling the 9 independent oysters in
`data/13-wgbs-samples.tsv` (duplicated libraries excluded; see README Known issues).

**Upstream:**

- `code/13-wgbs-submit.sh` (13.1 → 13.2 → 13.3) finished: reads the strand-merged
  CpG files `output/13-wgbs/methylation/*.CpG_report.merged_CpG_evidence.cov.gz`.
- `code/13.5-wgbs-gene-annotation.sh` finished: reads
  `output/13-wgbs/13.5-gene-annotation/GN_gene_annotation.tsv` (Swiss-Prot best hit +
  GO per GN gene). Only the last section needs it.

**Reference.** Chosen with the `ref` render parameter. `merged` (default) uses
`merged_out.fasta` + `GN.gene.gff3` and `output/13-wgbs/`. `hapa` uses the
`WGBS_REF=hapa` alignment to `Och_HapA_assembly.fa` (`output/13-wgbs-hapa/`), with the GN
gene models lifted onto HapA by `13.11` (same gene IDs, so `13.5` applies) and a HapA
SNP mask from `13.7` run with `WGBS_REF=hapa`; results go to
`output/13-wgbs-hapa/13.6-landscape`:
`rmarkdown::render("13.6-wgbs-methylation-landscape.Rmd", params = list(ref = "hapa"))`.
Either way the GFF has gene / mRNA / exon / CDS only, with every exon identical to a
CDS: there are **no UTRs**, so "exon" here means coding exon and 5'/3' UTR methylation
can't be separated from the flanks.

**Pooling.** At ~6–9x raw coverage per oyster, single-sample per-CpG estimates are
noisy. For the landscape, counts are summed across all 9 oysters per CpG (a
species-level methylome at roughly 9x the depth of one oyster). Per-oyster values are kept for
global, per-feature, and per-gene summaries, which pool many CpGs and so are stable
within a single animal; these show whether the patterns are consistent across
individuals and populations.

**Where to run:** a compute node inside the lab R container (~100 GB RAM), e.g.

```
salloc -A srlab -p cpu-g2-mem2x -c 8 --mem=150G -t 8:00:00
apptainer exec /gscratch/srlab/containers/srlab-R4.4-bioinformatics-container-c3d3116.sif R
```

Paths are relative to `code/`, like the other notebooks.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r packages}
# tidyverse first, so the Bioconductor set operations (reduce, setdiff, union, ...)
# mask the dplyr/purrr versions rather than the other way round.
library(tidyverse)
library(data.table)
library(GenomicRanges)
library(rtracklayer)
library(Biostrings)
library(Rsamtools)
```

```{r parameters}
ref <- match.arg(params$ref, c("merged", "hapa"))
ref_cfg <- list(
  merged = list(out_root = "../output/13-wgbs", genome_dir = "../data/13-wgbs-genome",
                fasta = "merged_out.fasta", gff = "GN.gene.gff3"),
  hapa   = list(out_root = "../output/13-wgbs-hapa", genome_dir = "../data/13-wgbs-genome-hapa",
                fasta = "Och_HapA_assembly.fa", gff = "GN.gene.hapa.gff3")   # gff from 13.11
)[[ref]]
out_root   <- ref_cfg$out_root
out_dir    <- file.path(out_root, "13.6-landscape")
meth_dir   <- file.path(out_root, "methylation")
genome_dir <- ref_cfg$genome_dir
annot_file <- "../output/13-wgbs/13.5-gene-annotation/GN_gene_annotation.tsv"
repeat_gff <- NULL   # RepeatMasker/EDTA GFF for this assembly, if one becomes available
snp_bed    <- file.path(out_root, "13.7-snps/wgbs_snps.q20.CpG.bed")   # from 13.7; skipped if absent
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# CpG level (pooled across oysters)
min_cov_pooled  <- 10     # summed reads per CpG
max_cov_pctl    <- 0.999  # drop pile-ups (collapsed repeats)
min_samples     <- 5L     # CpG must have data in >= 5 of 9 oysters
meth_breaks     <- c(-Inf, 10, 50, Inf)                 # % methylation
meth_labels     <- c("unmethylated (<10%)", "sparse (10-50%)", "methylated (>=50%)")

# features
flank_bp        <- 2000   # promoter (upstream) and downstream flank; same promoter as 13.4
window_size     <- 1e5    # chromosome-scale windows

# genes
min_gene_cpgs   <- 5      # covered CpGs needed for a gene-body estimate
min_gene_calls  <- 30     # per-oyster reads summed over the gene, for per-oyster values
n_body_bins     <- 20     # metagene: scaled gene body
n_flank_bins    <- 10     # metagene: each 2 kb flank

# function
min_go_genes    <- 10
max_go_genes    <- 500
```

# Samples

```{r samples}
samples <- read_tsv("../data/13-wgbs-samples.tsv", show_col_types = FALSE) %>%
  mutate(population = factor(population, levels = c("Pum", "Qui", "Rio")),
         file = file.path(meth_dir,
                          paste0(sample, ".CpG_report.merged_CpG_evidence.cov.gz")))
stopifnot(all(file.exists(samples$file)))
```

# Genome and features

Five non-overlapping compartments, so every base and CpG belongs to exactly one:

| feature | definition |
|---|---|
| exon | coding exon (no UTRs annotated) |
| intron | gene body minus exons |
| promoter | 2 kb upstream of a TSS, outside any gene |
| downstream | 2 kb downstream of a TES, outside any gene or promoter |
| intergenic | everything else |

```{r genome}
seqlens <- read_tsv(file.path(genome_dir, "genome.chrom.sizes"),
                    col_names = c("chr", "len"), show_col_types = FALSE)
si <- Seqinfo(seqlens$chr, seqlens$len)
fa <- FaFile(file.path(genome_dir, ref_cfg$fasta))

gff <- import(file.path(genome_dir, ref_cfg$gff))
seqlevels(gff, pruning.mode = "coarse") <- seqlevels(si)
seqinfo(gff) <- si

genes <- gff[gff$type == "gene"]
names(genes) <- genes$ID
exon_gr <- gff[gff$type == "exon"]
stopifnot(all(lengths(exon_gr$Parent) == 1))
exon_gr$tx <- as.character(unlist(exon_gr$Parent))
exon_gr$gene_id <- sub("\\.[0-9]+$", "", exon_gr$tx)

gene_red   <- reduce(genes, ignore.strand = TRUE)
exons      <- reduce(exon_gr, ignore.strand = TRUE)
introns    <- setdiff(gene_red, exons)
promoters  <- setdiff(reduce(trim(promoters(genes, upstream = flank_bp, downstream = 0)),
                             ignore.strand = TRUE), gene_red)
downstream <- setdiff(reduce(trim(flank(genes, flank_bp, start = FALSE)), ignore.strand = TRUE),
                      union(gene_red, promoters))
intergenic <- gaps(reduce(c(gene_red, promoters, downstream)))
intergenic <- intergenic[strand(intergenic) == "*"]

feature_levels <- c("exon", "intron", "promoter", "downstream", "intergenic")
features <- c(exons, introns, promoters, downstream, intergenic)
features$feature <- factor(rep(feature_levels,
                               lengths(list(exons, introns, promoters, downstream, intergenic))),
                           levels = feature_levels)
stopifnot(!any(countOverlaps(features, features) > 1))      # disjoint
stopifnot(sum(as.numeric(width(features))) == sum(as.numeric(seqlengths(si))))

annotate_feature <- function(gr) {
  features$feature[findOverlaps(gr, features, ignore.strand = TRUE, select = "first")]
}

# Every CpG dinucleotide in the assembly (1-based position of the C, as in Bismark).
genome_seq <- scanFa(fa)
names(genome_seq) <- sub(" .*", "", names(genome_seq))
genome_cpg <- do.call(c, lapply(seqlevels(si), function(chr) {
  GRanges(chr, IRanges(start(matchPattern("CG", genome_seq[[chr]])), width = 1),
          seqinfo = si)
}))
genome_cpg$feature <- annotate_feature(genome_cpg)
rm(genome_seq); invisible(gc())
message(length(genome_cpg), " CpGs in the assembly")
```

```{r feature-composition}
# Observed/expected CpG per feature: CG * length / (C * G). Germline methylation
# deaminates mC -> T over evolutionary time, so long-term methylated sequence is CpG-depleted.
composition <- map_dfr(feature_levels, function(f) {
  gr  <- features[features$feature == f]
  seq <- getSeq(fa, gr)
  bf  <- colSums(letterFrequency(seq, c("C", "G")))
  tibble(feature   = f,
         bp        = sum(as.numeric(width(gr))),
         n_cpg     = sum(as.numeric(vcountPattern("CG", seq))),
         gc_pct    = 100 * sum(bf) / bp,
         cpg_oe    = n_cpg * bp / (bf[["C"]] * bf[["G"]]))
}) %>%
  mutate(pct_genome = 100 * bp / sum(bp),
         cpg_per_kb = 1000 * n_cpg / bp)
composition
```

# Read CpG calls

Each oyster is read once. Per-oyster summaries (global, per feature, per gene) are
taken on the way through; per-CpG counts are summed into one pooled table.

**SNP mask.** CpGs carrying a SNP on their C or G (from `13.7`, biscuit genotyping of
the bisulfite BAMs) are dropped as each oyster is read, since a C→T SNP reads as an
unmethylated CpG. Every summary below is therefore SNP-masked; assembly CpG counts
(`genome_cpg`) are not. If the 13.7 file is missing, nothing is masked.

```{r read}
read_cov <- function(f) {
  fread(cmd = paste("gzip -dc", shQuote(f)), header = FALSE,
        select = c(1, 2, 5, 6), col.names = c("chr", "pos", "meth", "unmeth"))
}

snp_gr <- NULL
if (file.exists(snp_bed)) {
  snp_gr <- import(snp_bed, format = "bed")   # one base: the C or G hit
  message(length(snp_gr), " CpG-SNP sites from ", snp_bed)
} else {
  message("No SNP mask (", snp_bed, " not found); run code/13.7-wgbs-snps.sh to add one")
}

pooled <- NULL
ps_global <- ps_feature <- ps_gene <- vector("list", nrow(samples))
n_masked <- set_names(integer(nrow(samples)), samples$sample)

for (i in seq_len(nrow(samples))) {
  s <- samples$sample[i]
  d <- read_cov(samples$file[i])[chr %in% seqlevels(si)]
  if (!is.null(snp_gr)) {                     # each CpG spans its C (pos) and G (pos + 1)
    hit <- overlapsAny(GRanges(d$chr, IRanges(d$pos, width = 2)), snp_gr, ignore.strand = TRUE)
    n_masked[s] <- sum(hit)
    d <- d[!hit]
  }
  d[, cov := meth + unmeth]
  gr <- GRanges(d$chr, IRanges(d$pos, width = 1), seqinfo = si)
  d[, feature := annotate_feature(gr)]

  ps_global[[i]] <- d[, .(sample = s, cpgs_1x = .N, cpgs_5x = sum(cov >= 5),
                          median_cov = median(cov),
                          global_meth_pct = 100 * sum(meth) / sum(cov),
                          methylated_cpg_pct = 100 * mean(meth[cov >= 5] / cov[cov >= 5] >= 0.5))]
  ps_feature[[i]] <- d[, .(sample = s, meth = sum(meth), cov = sum(cov), n_cpg = .N),
                       by = feature]
  ov <- findOverlaps(gr, genes, ignore.strand = TRUE)
  ps_gene[[i]] <- data.table(gene_id = names(genes)[subjectHits(ov)],
                             meth = d$meth[queryHits(ov)], cov = d$cov[queryHits(ov)]
                             )[, .(sample = s, meth = sum(meth), cov = sum(cov), n_cpg = .N),
                               by = gene_id]

  pooled <- rbind(pooled, d[, .(chr, pos, meth, unmeth, n = 1L)]
                  )[, .(meth = sum(meth), unmeth = sum(unmeth), n = sum(n)), by = .(chr, pos)]
  message(s, ": ", nrow(d), " CpGs; pooled table now ", nrow(pooled))
}

ps_global  <- rbindlist(ps_global)  %>% left_join(samples, by = "sample")
ps_feature <- rbindlist(ps_feature) %>% left_join(samples, by = "sample")
ps_gene    <- rbindlist(ps_gene)
if (!is.null(snp_gr))
  write_tsv(tibble(sample = names(n_masked), cpgs_masked = n_masked),
            file.path(out_dir, "snp_mask_summary.tsv"))
rm(d, gr, ov); invisible(gc())
```

```{r pooled-filter}
pooled[, cov := meth + unmeth]
cov_max <- quantile(pooled$cov, max_cov_pctl)
pooled_f <- pooled[cov >= min_cov_pooled & cov <= cov_max & n >= min_samples]

cpg <- GRanges(pooled_f$chr, IRanges(pooled_f$pos, width = 1), seqinfo = si,
               meth = pooled_f$meth, cov = pooled_f$cov,
               pct = 100 * pooled_f$meth / pooled_f$cov)
cpg$feature <- annotate_feature(cpg)
cpg$class   <- cut(cpg$pct, meth_breaks, meth_labels, right = FALSE)

tibble(step = c("CpGs in assembly", "covered in >= 1 oyster", "pass pooled filters"),
       n = c(length(genome_cpg), nrow(pooled), length(cpg))) %>%
  mutate(pct_of_assembly = round(100 * n / length(genome_cpg), 1))

# Pooled per-CpG methylation track for IGV (alongside the 13.2 BAMs).
bw <- granges(cpg); bw$score <- cpg$pct
export.bw(sort(bw), file.path(out_dir, "pooled_CpG_methylation.bw"))
fwrite(pooled_f[, .(chr, start = pos, end = pos + 1L, pct = round(100 * meth / cov, 2),
                    meth, unmeth, n_samples = n)],
       file.path(out_dir, "pooled_CpG_methylation.tsv.gz"), sep = "\t")
rm(pooled, pooled_f); invisible(gc())
```

# Global methylation

## Per oyster

CpG methylation in invertebrates is moderate overall (bivalves typically ~10–20% of
CpGs methylated), unlike the ~70–80% of vertebrates.

```{r global-per-oyster}
aln <- read_tsv(file.path(out_root, "alignment_summary.tsv"), show_col_types = FALSE)
global <- ps_global %>%
  left_join(select(aln, sample, CHG_meth_pct, CHH_meth_pct), by = "sample") %>%
  select(sample, population, cpgs_1x, cpgs_5x, median_cov,
         global_meth_pct, methylated_cpg_pct, CHG_meth_pct, CHH_meth_pct)
write_tsv(global, file.path(out_dir, "global_methylation_per_oyster.tsv"))
global

global %>%
  pivot_longer(c(global_meth_pct, CHG_meth_pct, CHH_meth_pct),
               names_to = "context", values_to = "pct") %>%
  mutate(context = sub("_meth_pct", "", sub("global", "CpG", context))) %>%
  ggplot(aes(population, pct, colour = population)) +
  geom_jitter(width = 0.1, size = 2.5) +
  facet_wrap(~context, scales = "free_y") +
  labs(y = "% methylation (read-level)", title = "Methylation by context") +
  theme_bw() + theme(legend.position = "none")

kruskal.test(global_meth_pct ~ population, data = global)
```

## Per-CpG distribution (pooled)

A mosaic methylome shows up as a bimodal distribution: most CpGs near 0%, a smaller
fully methylated mode near 100%, and few intermediate sites.

```{r global-distribution}
ggplot(tibble(pct = cpg$pct), aes(pct)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  labs(x = "% methylation per CpG (pooled)", y = "CpGs") + theme_bw()

class_summary <- as_tibble(table(class = cpg$class)) %>%
  mutate(pct = 100 * n / sum(n))
write_tsv(class_summary, file.path(out_dir, "cpg_class_summary.tsv"))
class_summary
```

# Genomic features

```{r feature-summary}
cpg_tab <- as_tibble(as.data.frame(mcols(cpg)[, c("feature", "pct", "class")]))

feature_summary <- composition %>%
  left_join(as_tibble(table(feature = genome_cpg$feature)) %>% dplyr::rename(cpg_in_assembly = n),
            by = "feature") %>%
  left_join(cpg_tab %>%
              group_by(feature) %>%
              summarise(cpg_covered    = n(),
                        mean_meth      = mean(pct),
                        median_meth    = median(pct),
                        pct_methylated = 100 * mean(class == meth_labels[3]),
                        pct_sparse     = 100 * mean(class == meth_labels[2]),
                        pct_unmeth     = 100 * mean(class == meth_labels[1])),
            by = "feature") %>%
  mutate(feature = factor(feature, feature_levels),
         pct_cpg_covered = 100 * cpg_covered / cpg_in_assembly)
write_tsv(feature_summary, file.path(out_dir, "feature_summary.tsv"))
feature_summary
```

```{r feature-plots, fig.height=4}
cpg_tab %>%
  ggplot(aes(feature, fill = class)) +
  geom_bar(position = "fill") +
  scale_fill_manual(values = c("grey80", "#9ecae1", "#08519c")) +
  labs(y = "Proportion of covered CpGs", fill = NULL) + theme_bw()

cpg_tab %>%
  ggplot(aes(feature, pct)) +
  geom_violin(fill = "grey85", scale = "width") +
  stat_summary(fun = mean, geom = "point", colour = "red") +
  labs(y = "% methylation per CpG (pooled)") + theme_bw()
```

## Where the methylated CpGs are

Share of the genome, of all assembly CpGs, of covered CpGs, and of methylated CpGs
falling in each feature. Enrichment of methylated CpGs is tested per feature against
all other features (Fisher), using covered CpGs as the background.

```{r feature-enrichment}
meth_cpg <- cpg_tab$class == meth_labels[3]

feature_share <- bind_rows(
  composition %>% transmute(feature, set = "genome (bp)", n = bp),
  as_tibble(table(feature = genome_cpg$feature)) %>% mutate(set = "assembly CpGs"),
  as_tibble(table(feature = cpg_tab$feature)) %>% mutate(set = "covered CpGs"),
  as_tibble(table(feature = cpg_tab$feature[meth_cpg])) %>% mutate(set = "methylated CpGs")
) %>%
  group_by(set) %>% mutate(share = n / sum(n)) %>% ungroup() %>%
  mutate(set = fct_inorder(set), feature = factor(feature, feature_levels))

ggplot(feature_share, aes(set, share, fill = feature)) +
  geom_col() + scale_fill_brewer(palette = "Set2") +
  labs(x = NULL, y = "Share") + theme_bw()

feature_enrichment <- map_dfr(feature_levels, function(f) {
  in_f <- cpg_tab$feature == f
  ft <- fisher.test(table(in_f, meth_cpg))
  tibble(feature = f, odds_ratio = unname(ft$estimate),
         ci_low = ft$conf.int[1], ci_high = ft$conf.int[2], p = ft$p.value)
}) %>% mutate(p_adj = p.adjust(p, "BH"))
write_tsv(feature_enrichment, file.path(out_dir, "feature_methylated_cpg_enrichment.tsv"))
feature_enrichment
```

## Consistency across oysters

Per-oyster coverage-weighted methylation in each feature. If the feature ranking
holds in every animal, the pooled pattern isn't driven by a few deep samples.

```{r feature-per-oyster}
ps_feature %>%
  mutate(pct = 100 * meth / cov) %>%
  ggplot(aes(feature, pct, colour = population)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.6)) +
  labs(y = "% methylation (read-level, per oyster)") + theme_bw()
```

## Repeats

```{r repeats, eval = !is.null(repeat_gff)}
# Not run until a repeat annotation for merged_out.fasta exists. Repeats overlap the
# five features, so they are reported as a second, orthogonal layer.
reps <- import(repeat_gff)
seqlevels(reps, pruning.mode = "coarse") <- seqlevels(si)
rep_class <- sub("/.*", "", coalesce(reps$Target, reps$type))   # adjust to the GFF's class field
cpg$repeat_class <- "non-repeat"
h <- findOverlaps(cpg, reps, ignore.strand = TRUE, select = "first")
cpg$repeat_class[!is.na(h)] <- rep_class[h[!is.na(h)]]

as_tibble(as.data.frame(mcols(cpg)[, c("feature", "repeat_class", "pct")])) %>%
  group_by(feature, repeat_class) %>%
  summarise(n = n(), mean_meth = mean(pct), .groups = "drop") %>%
  arrange(feature, dplyr::desc(n))
```

# Chromosome-scale distribution

Methylation in `r window_size / 1000` kb windows along each chromosome, against gene
density. In mosaic methylomes methylation tracks gene density.

```{r windows, fig.width=11, fig.height=6}
win <- tileGenome(seqlengths(si), tilewidth = window_size, cut.last.tile.in.chrom = TRUE)
win <- binnedAverage(win, coverage(gene_red), "gene_frac")
win$n_cpg_assembly <- countOverlaps(win, genome_cpg)
h <- findOverlaps(cpg, win)
win_meth <- data.table(w = subjectHits(h), meth = cpg$meth[queryHits(h)],
                       cov = cpg$cov[queryHits(h)], pct = cpg$pct[queryHits(h)]
                       )[, .(meth_pct = 100 * sum(meth) / sum(cov),
                             frac_methylated = mean(pct >= 50), n_cpg = .N), by = w]

windows <- as_tibble(as.data.frame(win)) %>%
  mutate(w = row_number()) %>%
  left_join(win_meth, by = "w") %>%
  filter(n_cpg >= 20) %>%
  mutate(cpg_per_kb = 1000 * n_cpg_assembly / width)
write_tsv(windows, file.path(out_dir, "window_methylation.tsv"))

windows %>%
  transmute(seqnames, mb = (start + end) / 2e6,
            `CpG methylation (%)` = meth_pct, `gene coverage (%)` = 100 * gene_frac) %>%
  pivot_longer(-c(seqnames, mb)) %>%
  ggplot(aes(mb, value)) +
  geom_line(linewidth = 0.2) +
  geom_smooth(method = "loess", span = 0.2, se = FALSE, colour = "red", linewidth = 0.5) +
  facet_grid(name ~ seqnames, scales = "free", space = "free_x") +
  labs(x = "Position (Mb)", y = NULL) + theme_bw() +
  theme(strip.text.x = element_text(size = 7), axis.text.x = element_text(size = 6))

cor.test(windows$meth_pct, windows$gene_frac, method = "spearman")
cor.test(windows$meth_pct, windows$cpg_per_kb, method = "spearman")

ggplot(windows, aes(100 * gene_frac, meth_pct)) +
  geom_bin2d(bins = 60) + geom_smooth(method = "loess", colour = "red") +
  labs(x = "Gene coverage in window (%)", y = "Window CpG methylation (%)") + theme_bw()
```

# Gene-body methylation

## Per gene

Pooled coverage-weighted methylation over all covered CpGs in each gene, plus
the same for its exons, introns, and 2 kb promoter. Genes need ≥ `r min_gene_cpgs`
covered CpGs. CpG O/E and GC are computed from sequence for every gene.

```{r gene-level}
region_meth <- function(gr, ids, label) {
  h <- findOverlaps(cpg, gr, ignore.strand = TRUE)
  data.table(gene_id = ids[subjectHits(h)], meth = cpg$meth[queryHits(h)],
             cov = cpg$cov[queryHits(h)], pct = cpg$pct[queryHits(h)]
             )[, .(pct = 100 * sum(meth) / sum(cov), n = .N,
                   frac_meth_cpg = mean(pct >= 50)), by = gene_id] %>%
    as_tibble() %>%
    rename_with(~ paste0(label, "_", .x), -gene_id)
}

gene_promoters <- trim(promoters(genes, upstream = flank_bp, downstream = 0))
gene_introns   <- psetdiff(granges(genes),
                           split(granges(exon_gr), factor(exon_gr$gene_id, names(genes))))
gene_introns   <- unlist(gene_introns)

gene_seq  <- getSeq(fa, genes)
gene_comp <- letterFrequency(gene_seq, c("C", "G")) * 1   # double: C * G overflows integer
n_exons   <- table(factor(exon_gr$gene_id, names(genes)))

gene_meth <- tibble(gene_id   = names(genes),
                    chr       = as.character(seqnames(genes)),
                    length    = width(genes),
                    n_exons   = as.integer(n_exons[names(genes)]),
                    n_cpg     = as.numeric(vcountPattern("CG", gene_seq)),
                    gc_pct    = 100 * rowSums(gene_comp) / width(genes),
                    cpg_oe    = n_cpg * width(genes) / (gene_comp[, "C"] * gene_comp[, "G"])) %>%
  left_join(region_meth(genes, names(genes), "body"), by = "gene_id") %>%
  left_join(region_meth(exon_gr, exon_gr$gene_id, "exon"), by = "gene_id") %>%
  left_join(region_meth(gene_introns, names(gene_introns), "intron"), by = "gene_id") %>%
  left_join(region_meth(gene_promoters, names(gene_promoters), "promoter"), by = "gene_id") %>%
  mutate(body_n = replace_na(body_n, 0L),
         body_class = cut(if_else(body_n >= min_gene_cpgs, body_pct, NA_real_),
                          meth_breaks, meth_labels, right = FALSE))
rm(gene_seq); invisible(gc())

gbm <- filter(gene_meth, !is.na(body_class))
message(nrow(gbm), " of ", nrow(gene_meth), " genes with >= ", min_gene_cpgs, " covered CpGs")
table(gbm$body_class)
```

## Distribution

Bivalve gene-body methylation is usually bimodal: one group of genes heavily
methylated, the other nearly unmethylated. The local minimum of the density
between the modes is printed as a check on the fixed 10/50% class breaks.

```{r gene-distribution}
dens <- density(gbm$body_pct, from = 0, to = 100)
mid  <- dens$x > 10 & dens$x < 90
message("Density minimum between modes: ", round(dens$x[mid][which.min(dens$y[mid])], 1), "%")

ggplot(gbm, aes(body_pct)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  geom_vline(xintercept = c(10, 50), linetype = 2, colour = "grey50") +
  labs(x = "Gene-body CpG methylation (%)", y = "Genes") + theme_bw()

ggplot(gbm, aes(exon_pct, intron_pct)) +
  geom_bin2d(bins = 60) + geom_abline(linetype = 2, colour = "red") +
  labs(x = "Exon methylation (%)", y = "Intron methylation (%)") + theme_bw()

ggplot(gbm, aes(promoter_pct, body_pct)) +
  geom_bin2d(bins = 60) +
  labs(x = "Promoter (2 kb upstream) methylation (%)", y = "Gene-body methylation (%)") +
  theme_bw()
```

## CpG O/E and gene structure

Genes methylated in the germline lose CpGs over evolutionary time, so methylated
genes are expected to have lower CpG O/E. A bimodal CpG O/E distribution across all
genes (not only those with WGBS coverage) is the sequence-level signature of this.

```{r gene-structure}
ggplot(gene_meth, aes(cpg_oe)) +
  geom_histogram(bins = 80) + xlim(0, 1.5) +
  labs(x = "Gene CpG O/E (all genes)", y = "Genes") + theme_bw()

ggplot(gbm, aes(cpg_oe, fill = body_class)) +
  geom_density(alpha = 0.5) + xlim(0, 1.5) +
  scale_fill_manual(values = c("grey70", "#9ecae1", "#08519c")) +
  labs(x = "Gene CpG O/E", fill = NULL) + theme_bw()

cor.test(gbm$body_pct, gbm$cpg_oe, method = "spearman")

gbm %>%
  transmute(body_class, `gene length (kb)` = length / 1000, `exons` = n_exons,
            `GC (%)` = gc_pct) %>%
  pivot_longer(-body_class) %>%
  ggplot(aes(body_class, value, fill = body_class)) +
  geom_boxplot(outlier.size = 0.3) + scale_y_log10() +
  scale_fill_manual(values = c("grey70", "#9ecae1", "#08519c")) +
  facet_wrap(~name, scales = "free_y") + labs(x = NULL) + theme_bw() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 20, hjust = 1))

write_tsv(gene_meth, file.path(out_dir, "gene_methylation.tsv"))
```

## Metagene profile

Methylation along a scaled gene body (TSS → TES, `r n_body_bins` bins) with
`r flank_bp / 1000` kb flanks (`r n_flank_bins` bins each), oriented by strand. Flanks
can run into neighbouring genes; that is left in, as it is what a CpG near a TSS
actually sees.

```{r metagene, fig.height=4}
ext <- trim(resize(genes, width(genes) + 2 * flank_bp, fix = "center"))
h   <- findOverlaps(cpg, ext, ignore.strand = TRUE)
flank_step <- flank_bp / n_flank_bins

mg <- data.table(gene_id = names(genes)[subjectHits(h)],
                 pos = start(cpg)[queryHits(h)],
                 gs = start(genes)[subjectHits(h)], ge = end(genes)[subjectHits(h)],
                 minus = as.character(strand(genes))[subjectHits(h)] == "-",
                 meth = cpg$meth[queryHits(h)], cov = cpg$cov[queryHits(h)],
                 pct = cpg$pct[queryHits(h)])
mg[, `:=`(rel = fifelse(minus, ge - pos, pos - gs), len = ge - gs + 1)]   # 0 = TSS
mg[, bin := fcase(rel < 0,   floor(rel / flank_step),                       # -10 .. -1
                  rel < len, floor(rel / len * n_body_bins),                #   0 .. 19
                  default = n_body_bins + floor((rel - len) / flank_step))] #  20 .. 29
mg <- mg[bin >= -n_flank_bins & bin < n_body_bins + n_flank_bins]
mg <- merge(mg, as.data.table(gene_meth[, c("gene_id", "body_class")]), by = "gene_id")

metagene <- rbind(
  mg[, .(body_class = "all genes", meth_pct = 100 * sum(meth) / sum(cov),
         frac_methylated = mean(pct >= 50), n_cpg = .N), by = bin],
  mg[!is.na(body_class), .(meth_pct = 100 * sum(meth) / sum(cov),
                           frac_methylated = mean(pct >= 50), n_cpg = .N),
     by = .(bin, body_class = as.character(body_class))],
  use.names = TRUE)
metagene[, body_class := factor(body_class, c("all genes", meth_labels))]
write_tsv(metagene, file.path(out_dir, "metagene_profile.tsv"))

ggplot(metagene, aes(bin + 0.5, meth_pct, colour = body_class)) +
  geom_line() +
  geom_vline(xintercept = c(0, n_body_bins), linetype = 2, colour = "grey50") +
  scale_x_continuous(breaks = c(-n_flank_bins, 0, n_body_bins, n_body_bins + n_flank_bins),
                     labels = c(paste0("-", flank_bp / 1000, " kb"), "TSS", "TES",
                                paste0("+", flank_bp / 1000, " kb"))) +
  scale_colour_manual(values = c("black", "grey60", "#9ecae1", "#08519c")) +
  labs(x = NULL, y = "CpG methylation (%)", colour = NULL) + theme_bw()
rm(mg); invisible(gc())
```

## Exon and intron rank

Methylation by position of the exon/intron along the transcript (5' → 3'), for
multi-exon genes. In many invertebrates the first exon and 5' end are less
methylated than the rest of the body.

```{r exon-rank, fig.height=4}
ex <- data.table(seqnames = as.character(seqnames(exon_gr)), start = start(exon_gr),
                 end = end(exon_gr), strand = as.character(strand(exon_gr)),
                 tx = exon_gr$tx, gene_id = exon_gr$gene_id)
setorder(ex, tx, start)
ex[, `:=`(n_ex = .N, idx = seq_len(.N)), by = tx]
ex[, rank := fifelse(strand == "-", n_ex - idx + 1L, idx)]

intr <- ex[n_ex > 1, .(seqnames = seqnames[1], start = end[-.N] + 1L, end = start[-1] - 1L,
                       strand = strand[1], n_ex = n_ex[1], idx = seq_len(.N - 1)), by = tx]
intr <- intr[end >= start]
intr[, rank := fifelse(strand == "-", n_ex - idx, idx)]

# Ranks 1-5, then "6+"; the final element of each transcript is pulled out as "last"
# (except a lone intron, which stays rank 1).
rank_meth <- function(dt, what) {
  n_el  <- if (what == "exon") dt$n_ex else dt$n_ex - 1L
  label <- fifelse(dt$rank == n_el & dt$rank > 1, "last", as.character(pmin(dt$rank, 6)))
  gr <- makeGRangesFromDataFrame(dt, seqinfo = si)
  h  <- findOverlaps(cpg, gr, ignore.strand = TRUE)
  data.table(rank_label = label[subjectHits(h)],
             meth = cpg$meth[queryHits(h)], cov = cpg$cov[queryHits(h)]
             )[, .(meth_pct = 100 * sum(meth) / sum(cov), n_cpg = .N), by = rank_label
             ][, element := what][]
}

rank_tab <- rbind(rank_meth(ex[n_ex > 1], "exon"), rank_meth(intr, "intron")) %>%
  mutate(rank_label = factor(recode(rank_label, `6` = "6+"),
                             c(as.character(1:5), "6+", "last")))
write_tsv(rank_tab, file.path(out_dir, "exon_intron_rank_methylation.tsv"))

ggplot(rank_tab, aes(rank_label, meth_pct, colour = element, group = element)) +
  geom_point(size = 2) + geom_line() +
  labs(x = "Rank from 5' end", y = "CpG methylation (%)", colour = NULL,
       subtitle = "Multi-exon genes; 'last' = final exon / intron") + theme_bw()
```

## Variability among oysters

Per-oyster gene-body methylation (reads summed over each gene) for genes with
≥ `r min_gene_calls` calls in every oyster. In bivalves, heavily methylated genes
tend to be the most stable among individuals.

```{r gene-variability}
gene_ind <- ps_gene[cov >= min_gene_calls & n_cpg >= min_gene_cpgs]
gene_ind <- gene_ind[, if (.N == nrow(samples)) .SD, by = gene_id]
gene_ind[, pct := 100 * meth / cov]

gene_var <- gene_ind[, .(mean_pct = mean(pct), sd_pct = sd(pct)), by = gene_id] %>%
  as_tibble() %>%
  left_join(select(gene_meth, gene_id, body_class), by = "gene_id")
write_tsv(gene_var, file.path(out_dir, "gene_methylation_variability.tsv"))
message(nrow(gene_var), " genes with per-oyster estimates in all ", nrow(samples), " oysters")

ggplot(gene_var, aes(mean_pct, sd_pct)) +
  geom_bin2d(bins = 60) + geom_smooth(method = "loess", colour = "red") +
  labs(x = "Mean gene-body methylation across oysters (%)",
       y = "SD among oysters (% points)") + theme_bw()

# Between-oyster agreement in gene-body methylation (pairwise Spearman).
pm_gene <- dcast(gene_ind, gene_id ~ sample, value.var = "pct")
cc <- cor(as.matrix(pm_gene[, -1]), method = "spearman")
summary(cc[upper.tri(cc)])
```

# Methylation and gene function

Functional annotation from `13.5` (DIAMOND blastp of GN proteins vs Swiss-Prot,
E ≤ 1e-8, GO from the best hit). GO terms are the direct UniProt annotations, not
propagated to parent terms, so broad terms are under-counted; treat the term tests
as descriptive.

```{r load-annotation}
annot <- read_tsv(annot_file, show_col_types = FALSE,
                  col_types = cols(.default = col_character()))

gene_fun <- gene_meth %>%
  left_join(select(annot, gene_id, accession, protein_name, go_bp), by = "gene_id") %>%
  mutate(annotation = case_when(is.na(accession) ~ "no Swiss-Prot hit",
                                is.na(go_bp)     ~ "hit, no GO BP",
                                TRUE             ~ "hit with GO BP"),
         annotation = factor(annotation, c("hit with GO BP", "hit, no GO BP",
                                           "no Swiss-Prot hit")))
write_tsv(gene_fun, file.path(out_dir, "gene_methylation_annotated.tsv"))
```

## Annotated vs unannotated genes

Genes with a Swiss-Prot hit are mostly conserved; unannotated genes are enriched for
lineage-specific and mis-predicted genes. Conserved genes are expected to carry more
gene-body methylation.

```{r annotation-status}
gf <- filter(gene_fun, !is.na(body_class))

gf %>% count(annotation, body_class) %>%
  group_by(annotation) %>% mutate(pct = 100 * n / sum(n)) %>% ungroup()
chisq.test(table(gf$annotation, gf$body_class))

ggplot(gf, aes(annotation, body_pct, fill = annotation)) +
  geom_violin(scale = "width") + geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  labs(x = NULL, y = "Gene-body methylation (%)") + theme_bw() +
  theme(legend.position = "none")
```

## GO term tests

Background: genes with a gene-body estimate and ≥ 1 GO BP term. Two tests per term
(`r min_go_genes`–`r max_go_genes` genes):

- **Wilcoxon rank-sum** of gene-body methylation, term genes vs the rest of the
  background (a continuous, direction-aware test; `delta_median` > 0 = term genes more
  methylated).
- **Fisher** for over-representation among methylated (≥ 50%) and among unmethylated
  (< 10%) genes.

```{r go-tests}
go_long <- gf %>%
  filter(!is.na(go_bp)) %>%
  select(gene_id, body_pct, body_class, go_bp) %>%
  separate_rows(go_bp, sep = "; ") %>%
  mutate(go_id = str_extract(go_bp, "GO:\\d{7}"),
         term  = str_trim(str_remove(go_bp, "\\s*\\[GO:\\d{7}\\]$"))) %>%
  filter(!is.na(go_id)) %>%
  distinct(gene_id, go_id, .keep_all = TRUE)

bg <- distinct(go_long, gene_id, body_pct, body_class)
is_high <- set_names(bg$body_class == meth_labels[3], bg$gene_id)
is_low  <- set_names(bg$body_class == meth_labels[1], bg$gene_id)

terms <- go_long %>% count(go_id, term) %>%
  filter(n >= min_go_genes, n <= max_go_genes)

go_tests <- map_dfr(seq_len(nrow(terms)), function(k) {
  ids <- go_long$gene_id[go_long$go_id == terms$go_id[k]]
  in_t <- bg$gene_id %in% ids
  wt <- wilcox.test(bg$body_pct[in_t], bg$body_pct[!in_t])
  fisher_p <- function(flag) fisher.test(table(factor(in_t, c(TRUE, FALSE)),
                                               factor(flag, c(TRUE, FALSE))),
                                         alternative = "greater")$p.value
  tibble(go_id = terms$go_id[k], term = terms$term[k], n_genes = sum(in_t),
         median_meth = median(bg$body_pct[in_t]),
         delta_median = median(bg$body_pct[in_t]) - median(bg$body_pct[!in_t]),
         pct_methylated_genes = 100 * mean(is_high[in_t]),
         wilcox_p = wt$p.value,
         high_fisher_p = fisher_p(is_high),
         low_fisher_p  = fisher_p(is_low))
}) %>%
  mutate(across(ends_with("_p"), ~ p.adjust(.x, "BH"), .names = "{.col}_adj")) %>%
  arrange(wilcox_p)

write_tsv(go_tests, file.path(out_dir, "GO_BP_methylation_tests.tsv"))
message(nrow(bg), " background genes; ", nrow(terms), " GO BP terms tested; overall ",
        round(100 * mean(is_high), 1), "% of background genes methylated")

go_tests %>% filter(wilcox_p_adj < 0.05) %>% count(direction = sign(delta_median))
```

```{r go-plot, fig.height=8, fig.width=9}
top_go <- go_tests %>%
  filter(wilcox_p_adj < 0.05) %>%
  group_by(direction = if_else(delta_median > 0, "higher methylation", "lower methylation")) %>%
  slice_min(wilcox_p, n = 15, with_ties = FALSE) %>% ungroup()

if (nrow(top_go) > 0) {
  top_go %>%
    mutate(term = fct_reorder(str_trunc(term, 60), median_meth)) %>%
    ggplot(aes(median_meth, term, size = n_genes, colour = -log10(wilcox_p_adj))) +
    geom_point() +
    geom_vline(xintercept = median(bg$body_pct), linetype = 2) +
    facet_grid(direction ~ ., scales = "free_y", space = "free_y") +
    scale_colour_viridis_c() +
    labs(x = "Median gene-body methylation of term genes (%)", y = NULL,
         subtitle = "Dashed line: background median") + theme_bw()
}
```

## Broad functional groups

A coarse view, grouping genes by keywords in their GO BP term names. The groups
follow the housekeeping-vs-inducible contrast often reported for bivalve gene-body
methylation (constitutively expressed, housekeeping functions methylated;
environmentally responsive functions less so). Genes can fall into more than one
group; edit the patterns to suit.

```{r functional-groups, fig.height=5}
func_groups <- c(
  "Translation / ribosome"        = "translation|ribosom|rRNA processing",
  "RNA processing / splicing"     = "mRNA processing|RNA splicing|spliceosom",
  "DNA replication / repair"      = "DNA replication|DNA repair",
  "Cell cycle"                    = "cell cycle|cell division|mitotic",
  "Mitochondria / energy"         = "mitochondri|electron transport|tricarboxylic|ATP synthesis",
  "Protein degradation"           = "ubiquitin|proteasom|protein catabolic",
  "Signal transduction"           = "signal transduction|signaling pathway|G protein-coupled",
  "Immune / defense"              = "immune|defense response|inflammatory|innate",
  "Stress response"               = "response to stress|oxidative stress|heat|hypoxia|unfolded protein",
  "Development / differentiation" = "development|morphogenesis|differentiation")

group_meth <- imap_dfr(func_groups, function(pat, grp) {
  go_long %>% filter(str_detect(term, regex(pat, ignore_case = TRUE))) %>%
    distinct(gene_id, body_pct) %>% mutate(group = grp)
})

group_stats <- group_meth %>%
  group_by(group) %>%
  summarise(n_genes = n(), median_meth = median(body_pct),
            pct_methylated_genes = 100 * mean(body_pct >= 50),
            wilcox_p = wilcox.test(body_pct, bg$body_pct[!bg$gene_id %in% gene_id])$p.value) %>%
  mutate(wilcox_p_adj = p.adjust(wilcox_p, "BH")) %>%
  arrange(dplyr::desc(median_meth))
write_tsv(group_stats, file.path(out_dir, "functional_group_methylation.tsv"))
group_stats

group_meth %>%
  mutate(group = factor(group, group_stats$group)) %>%
  ggplot(aes(body_pct, group)) +
  geom_boxplot(outlier.size = 0.3) +
  geom_vline(xintercept = median(bg$body_pct), linetype = 2) +
  labs(x = "Gene-body methylation (%)", y = NULL,
       subtitle = "Dashed line: median of all genes with GO BP") + theme_bw()
```

# Summary

```{r summary}
summary_tab <- tribble(
  ~metric, ~value,
  "CpGs in assembly",                         length(genome_cpg),
  "CpGs passing pooled filters",              length(cpg),
  "mean global CpG methylation, per oyster (%)", mean(global$global_meth_pct),
  "methylated CpGs (>=50%, pooled) (%)",      100 * mean(cpg$pct >= 50),
  "genes with gene-body estimate",            nrow(gbm),
  "methylated genes (>=50%) (%)",             100 * mean(gbm$body_class == meth_labels[3]),
  "unmethylated genes (<10%) (%)",            100 * mean(gbm$body_class == meth_labels[1]),
  "rho gene methylation vs CpG O/E",          cor(gbm$body_pct, gbm$cpg_oe, method = "spearman",
                                                  use = "complete.obs"),
  "rho window methylation vs gene coverage",  cor(windows$meth_pct, windows$gene_frac,
                                                  method = "spearman"),
  "GO BP terms tested",                       nrow(go_tests),
  "GO BP terms, Wilcoxon FDR < 0.05",         sum(go_tests$wilcox_p_adj < 0.05)
) %>% mutate(value = signif(value, 4))
write_tsv(summary_tab, file.path(out_dir, "landscape_summary.tsv"))
summary_tab
```

# Next steps

- **Repeats / TEs.** 69% of the assembly is repeat. Run RepeatMasker (or EDTA) on
  `merged_out.fasta`, set `repeat_gff`, and the repeat chunk will split each feature
  into repeat vs non-repeat; intronic TEs are a common source of gene-body
  methylation.
- **Expression.** The classic test of the gene-body model is methylation vs
  expression level and breadth. Join `gene_methylation.tsv` to per-gene expression
  from `09` (mean and CV of normalized counts across the 12 RNA-seq oysters) on GN
  gene ID. These are different animals from the WGBS oysters, so this is a
  gene-level, not individual-level, comparison.
- **GO propagation.** With `GO.db` (or `topGO`), propagate annotations to ancestors
  before the term tests; that will mainly add power to the broad terms.
- **SNP filter.** C/T SNPs at CpGs read as unmethylated. For the landscape this
  mainly lowers the methylated fraction slightly; it matters more for `13.4`.

```{r session}
sessionInfo()
```
