---
title: "13.8 - GO enrichment of DMR genes"
output:
  html_document:
    toc: true
    toc_float: true
date: "2026-10-10"
params:
  dmr_class: "all"   # or a 13.9 class, e.g. "no genetic signal"
---

Are genes near population differentially methylated regions (DMRs) enriched for
particular functions? DMRs come from `13.4` (1 kb tiles, SNP-masked); GO terms come
from the Swiss-Prot best hit of each GN gene (`13.5`).

**Gene sets.**

- *DMR genes:* genes whose body or `promoter_bp` upstream overlaps a DMR, the same
  promoter definition as the `13.4` feature annotation.
- *Background:* genes whose body or promoter overlaps a tile that 13.4 actually
  tested (`tiles_tested.bed`), so the comparison is against genes that *could* have
  produced a DMR, not the whole genome. Both sets are restricted to genes with at
  least one GO term.

**Test.** One-sided Fisher's exact test per GO term, with each gene's terms
propagated to all ancestor terms (the topGO "classic" method, written out here
because topGO is not in the lab container). Terms with fewer than `min_term` or more
than `max_term` background genes are skipped; BH correction is per ontology.

BH runs over every term in the size range; only terms with at least `min_hits` DMR
genes are then reported.

**Caveats.** Most DMRs are intergenic, so the DMR gene sets are small and the test
has little power. GO terms come from each gene's best Swiss-Prot hit, so tandem
paralogs that share a hit (and often sit in the same DMR) count as independent genes
and can make one locus look like an enriched term; check the `genes` column. Treat
individual terms as leads, not results.

**Restricting to clean DMRs.** Render with `params = list(dmr_class = "no genetic
signal")` to keep only DMRs that the `13.9` bisulfite SNP check found free of
population-differentiated variants, CpG-destroying SNPs and repeat signal. Other
classes (`untested`, `repeat-like`, `genetic: CpG-destroying`, `genetic: other`) work
the same way. Results then go to `13.8-dmr-go.<class>/` so the all-DMR run is kept.
The background is unchanged (all genes on tested tiles); since a DMR can only be
classed clean where every oyster has enough depth to test for SNPs, the clean set
leans toward well-covered regions, which the background does not.

**Upstream:** `13.4` (with `tiles_tested.bed`) and `13.5`; `13.9` when `dmr_class`
is set. Paths are relative to
`code/`, like the other notebooks; runs in the lab R container.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6)
```

```{r packages}
library(GenomicRanges)
library(rtracklayer)
library(GO.db)
library(tidyverse)
# tidyverse masks the range versions of reduce()/setdiff(); none are used here.
```

```{r parameters}
dm_dir      <- "../output/13-wgbs/13.4-methylation"
annot_file  <- "../output/13-wgbs/13.5-gene-annotation/GN_gene_annotation.tsv"
genome_dir  <- "../data/13-wgbs-genome"
snp_dir     <- "../output/13-wgbs/13.9-bs-snp-check"   # batch_DMR_<contrast>.tsv
dmr_class   <- params$dmr_class                         # "all" = no 13.9 filter
out_dir     <- paste0("../output/13-wgbs/13.8-dmr-go",
                      if (dmr_class == "all") "" else
                        paste0(".", gsub("[^A-Za-z0-9]+", "-", dmr_class)))
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

promoter_bp <- 2000
min_term    <- 5      # background genes per term
max_term    <- 500
min_hits    <- 2      # DMR genes per term to report
fdr         <- 0.05
```

# Genes, DMRs, tested tiles

```{r inputs}
seqlens <- read_tsv(file.path(genome_dir, "genome.chrom.sizes"),
                    col_names = c("chr", "len"), show_col_types = FALSE)
si <- Seqinfo(seqlens$chr, seqlens$len)

gff <- import(file.path(genome_dir, "GN.gene.gff3"))
seqlevels(gff, pruning.mode = "coarse") <- seqlevels(si)
seqinfo(gff) <- si
genes <- gff[gff$type == "gene"]
names(genes) <- genes$ID

# Gene body plus upstream promoter, strand-aware.
gene_ext <- trim(resize(genes, width(genes) + promoter_bp, fix = "end"))

read_dm <- function(f) {
  read_tsv(f, show_col_types = FALSE) %>%
    { GRanges(.$seqnames, IRanges(.$start, .$end), meth.diff = .$meth.diff,
              feature = .$feature) }
}
dmr_files <- list.files(dm_dir, pattern = "^DMR_.*_vs_.*\\.tsv$", full.names = TRUE)
dmrs <- set_names(map(dmr_files, read_dm),
                  sub("^DMR_(.*)\\.tsv$", "\\1", basename(dmr_files)))
tiles <- import(file.path(dm_dir, "tiles_tested.bed"))

# 13.9 class per DMR (matched on coordinates), and the optional class filter.
dmrs <- imap(dmrs, function(d, cn) {
  f <- file.path(snp_dir, paste0("batch_DMR_", cn, ".tsv"))
  if (!file.exists(f)) {
    if (dmr_class != "all") stop("dmr_class is set but ", f, " is missing; run 13.9 first")
    d$snp_class <- NA_character_
    return(d)
  }
  cls <- read_tsv(f, show_col_types = FALSE) %>%
    transmute(key = paste(seqnames, start, end), class)
  d$snp_class <- cls$class[match(paste(seqnames(d), start(d), end(d)), cls$key)]
  stopifnot(!anyNA(d$snp_class))      # 13.9 must have been run on these DMRs
  if (dmr_class == "all") d else d[d$snp_class == dmr_class]
})
message("DMR class filter: ", dmr_class)

map_int(dmrs, length)
length(tiles)
```

# GO annotation

`go_ids` holds the Swiss-Prot best hit's GO terms. Each gene gets those terms plus
all their ancestors in the same ontology.

```{r go}
annot <- read_tsv(annot_file, show_col_types = FALSE)

gene_go <- annot %>%
  select(gene_id, go_ids) %>%
  filter(!is.na(go_ids)) %>%
  separate_rows(go_ids, sep = ";\\s*") %>%
  rename(go_id = go_ids) %>%
  filter(str_detect(go_id, "^GO:\\d{7}$")) %>%
  distinct()

go_ont <- suppressMessages(AnnotationDbi::select(
  GO.db, keys = unique(gene_go$go_id), columns = c("ONTOLOGY", "TERM"), keytype = "GOID"))

ancestors <- list(BP = as.list(GOBPANCESTOR), MF = as.list(GOMFANCESTOR),
                  CC = as.list(GOCCANCESTOR))

gene_go_prop <- gene_go %>%
  inner_join(go_ont, by = c("go_id" = "GOID")) %>%
  mutate(anc = map2(go_id, ONTOLOGY, ~ c(.x, setdiff(ancestors[[.y]][[.x]], "all")))) %>%
  select(gene_id, ontology = ONTOLOGY, anc) %>%
  unnest(anc) %>%
  rename(go_id = anc) %>%
  distinct()

term_names <- suppressMessages(AnnotationDbi::select(
  GO.db, keys = unique(gene_go_prop$go_id), columns = "TERM", keytype = "GOID")) %>%
  rename(go_id = GOID, term = TERM)

tibble(genes_with_go = n_distinct(gene_go$gene_id),
       direct_terms  = n_distinct(gene_go$go_id),
       terms_after_propagation = n_distinct(gene_go_prop$go_id))
```

# Enrichment

```{r enrichment}
genes_hit <- function(gr) names(genes)[unique(subjectHits(findOverlaps(gr, gene_ext, ignore.strand = TRUE)))]

universe <- intersect(genes_hit(tiles), unique(gene_go_prop$gene_id))

go_fisher <- function(set, universe) {
  set <- intersect(set, universe)
  gp  <- filter(gene_go_prop, gene_id %in% universe)
  gp %>%
    group_by(ontology, go_id) %>%
    summarise(bg = n(), hits = sum(gene_id %in% set),
              genes = paste(sort(gene_id[gene_id %in% set]), collapse = ";"),
              .groups = "drop") %>%
    filter(bg >= min_term, bg <= max_term) %>%
    mutate(n_set = length(set), n_universe = length(universe),
           expected = bg * n_set / n_universe,
           fold = hits / expected,
           p = map2_dbl(hits, bg, ~ fisher.test(
             matrix(c(.x, length(set) - .x, .y - .x,
                      length(universe) - length(set) - .y + .x), 2),
             alternative = "greater")$p.value)) %>%
    group_by(ontology) %>%
    mutate(fdr = p.adjust(p, "BH")) %>%   # over every term in the size range,
    ungroup() %>%                          # before dropping low-hit terms
    filter(hits >= min_hits) %>%
    left_join(term_names, by = "go_id") %>%
    arrange(ontology, p) %>%
    select(ontology, go_id, term, hits, bg, expected, fold, p, fdr, genes)
}

results <- list()
gene_sets <- list()
for (cn in names(dmrs)) {
  d <- dmrs[[cn]]
  sets <- list(all = genes_hit(d),
               hyper = genes_hit(d[d$meth.diff > 0]),   # higher in the 2nd population
               hypo  = genes_hit(d[d$meth.diff < 0]))
  for (dir in names(sets)) {
    key <- paste(cn, dir, sep = "_")
    gene_sets[[key]] <- sets[[dir]]
    results[[key]] <- go_fisher(sets[[dir]], universe)
    write_tsv(results[[key]], file.path(out_dir, paste0("GO_", key, ".tsv")))
  }
}

summary_tbl <- imap_dfr(gene_sets, ~ tibble(
  set = .y,
  dmr_genes = length(.x),
  dmr_genes_with_go = length(intersect(.x, universe)),
  universe = length(universe),
  terms_reported = nrow(results[[.y]]),
  terms_fdr = sum(results[[.y]]$fdr < fdr)))
write_tsv(summary_tbl, file.path(out_dir, "GO_summary.tsv"))
summary_tbl
```

## Top terms

```{r top-terms}
top <- imap_dfr(results, ~ .x %>% group_by(ontology) %>% slice_min(p, n = 10, with_ties = FALSE) %>%
                  ungroup() %>% mutate(set = .y)) %>%
  filter(str_ends(set, "_all"))
top %>% select(set, ontology, term, hits, bg, fold, p, fdr)

if (nrow(top) > 0) {
  top %>%
    mutate(term = fct_reorder(str_trunc(term, 60), -log10(p))) %>%
    ggplot(aes(-log10(p), term, size = hits, colour = fdr < fdr)) +
    geom_point() +
    facet_grid(ontology ~ ., scales = "free_y", space = "free_y") +
    scale_colour_manual(values = c(`FALSE` = "grey50", `TRUE` = "firebrick"),
                        name = paste("FDR <", fdr)) +
    labs(x = "-log10 Fisher p", y = NULL,
         title = "GO terms among DMR genes (body + promoter)") +
    theme_bw()
}
```

## DMR genes

Every gene near a DMR with its best Swiss-Prot hit, for reading through directly.

```{r dmr-genes}
dmr_gene_tbl <- imap_dfr(dmrs, function(d, cn) {
  ov <- findOverlaps(d, gene_ext, ignore.strand = TRUE)
  tibble(contrast = cn, gene_id = names(genes)[subjectHits(ov)],
         dmr = paste0(seqnames(d), ":", start(d), "-", end(d))[queryHits(ov)],
         meth_diff = d$meth.diff[queryHits(ov)],
         dmr_feature = as.character(d$feature[queryHits(ov)]),
         snp_class = d$snp_class[queryHits(ov)])
}) %>%
  left_join(select(annot, gene_id, entry_name, protein_name, evalue), by = "gene_id") %>%
  arrange(contrast, desc(abs(meth_diff)))
write_tsv(dmr_gene_tbl, file.path(out_dir, "DMR_genes.tsv"))
dmr_gene_tbl %>% select(gene_id, meth_diff, dmr_feature, snp_class, protein_name)
```

```{r session}
sessionInfo()
```
