---
title: "14.2 - DNA methylation: Ostrea chilensis vs Ostrea lurida"
output:
  html_document:
    toc: true
    toc_float: true
date: "2026-10-08"
---

Comparative DNA methylation between the Chilean flat oyster (*O. chilensis*, Och)
and the Olympia oyster (*O. lurida*, Olur), at three levels:

1. **Methylome landscape** — global CpG methylation, the per-CpG distribution, and
   how methylation is partitioned among exons, introns, promoters and intergenic DNA.
2. **Gene-body methylation (gbM)** — per-gene levels and the methylated / unmethylated
   gene classes typical of bivalves.
3. **Conservation across orthologs** — do one-to-one orthologs keep their methylation
   level and class between the two species? Genomic CpG observed/expected (CpG o/e),
   the long-term footprint of germline methylation, is compared alongside the
   measured levels and is available even for a species without methylation data.

**Inputs** (all listed in `data/14-species.tsv`, paths relative to the repo root):

| | *O. chilensis* | *O. lurida* |
|---|---|---|
| genome + GFF | `merged_out.fasta`, `GN.gene.gff3` (as in `09`, `13`) | set in `data/14-species.tsv` |
| per-CpG calls | 15 WGBS oysters from `13.2` | `data/14-Olur-meth-samples.tsv` (to fill in) |
| orthologs | `output/14-meth-compare/orthologs/Och_Olur_rbh.tsv` from `14.1` | |

Per-CpG files are Bismark coverage format, ideally strand-merged
(`coverage2cytosine --merge_CpG`). The *O. lurida* calls must be on the same assembly
as its GFF: published Olympia oyster methylation data were aligned to `Olurida_v081`,
not the newer chromosome-level `xbOstLuri2` assembly, so either use the matching
v081 annotation or re-align the reads.

**Assay caveat.** If the *O. lurida* data are MBD-BS or RRBS rather than WGBS, the
enrichment step inflates absolute methylation and drops unmethylated regions. In
that case read the species differences in global level and feature partitioning as
qualitative, and lean on within-species ranks (the gene class and Spearman
correlations below), which are robust to the shift.

**Upstream:** `13` finished for *O. chilensis*; `code/14.1-meth-compare-orthologs.sh` run.
Run where `13.4` runs (lab R container on a compute node, ~100 GB RAM for the
pooled CpG tables). Paths are relative to `code/`, like the other notebooks.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 5)
```

```{r packages}
library(data.table)
library(GenomicRanges)
library(rtracklayer)
library(Rsamtools)
library(Biostrings)
library(tidyverse)
# tidyverse masks the range versions of reduce() (purrr) and setdiff() (dplyr),
# so range operations below are namespaced.
```

```{r parameters}
out_dir <- "../output/14-meth-compare"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

sample_min_cov <- 5      # per-sample CpG depth for per-oyster summaries
pooled_min_cov <- 10     # species-level CpG depth after summing oysters
min_gene_cpgs  <- 5      # covered CpGs required per gene for gbM
promoter_bp    <- 2000   # upstream window treated as promoter
meth_hi        <- 50     # % at or above: methylated CpG
meth_lo        <- 10     # % below: unmethylated CpG (between = sparse)
```

# Species and samples

```{r species}
species <- read_tsv("../data/14-species.tsv", show_col_types = FALSE) %>%
  mutate(across(c(genome_fa, gff, meth_samples), ~ file.path("..", .x)))
stopifnot(nrow(species) == 2)
sp_keys   <- species$species                     # c("Och", "Olur")
sp_labels <- set_names(species$label, sp_keys)

samples <- map_dfr(seq_len(nrow(species)), function(i) {
  f <- species$meth_samples[i]
  if (!file.exists(f)) return(NULL)
  read_tsv(f, comment = "#", show_col_types = FALSE,
           col_types = cols(.default = "c")) %>%
    mutate(species = species$species[i], cov_file = file.path("..", cov_file))
})

missing <- samples %>% filter(!file.exists(cov_file))
if (nrow(missing) > 0) {
  warning(nrow(missing), " CpG files not found; dropping them:\n",
          paste(missing$cov_file, collapse = "\n"))
  samples <- samples %>% filter(file.exists(cov_file))
}

has_meth <- set_names(sp_keys %in% samples$species, sp_keys)
if (!any(has_meth)) message("No CpG files for either species: running CpG o/e sections only.")
samples %>% count(species, assay, group)
has_meth
```

`has_meth` controls which sections run per species: a species without CpG files is
still carried through the CpG o/e and ortholog sections. With no CpG files at all,
the measured-methylation sections are skipped.

# Genome annotation

The same feature definitions as `13.4`, applied to each genome: protein-coding genes
are those with an mRNA child, and each CpG is assigned to one feature by priority
exon > intron > promoter > intergenic.

```{r features}
build_annotation <- function(genome_fa, gff_file) {
  if (!file.exists(paste0(genome_fa, ".fai"))) indexFa(genome_fa)
  fai <- read_tsv(paste0(genome_fa, ".fai"), col_names = c("chr", "len"),
                  col_select = 1:2, show_col_types = FALSE)
  si  <- Seqinfo(fai$chr, fai$len)

  gff <- import(gff_file)
  seqlevels(gff, pruning.mode = "coarse") <- seqlevels(si)   # align to FASTA order
  seqinfo(gff) <- si
  first_parent <- function(x) as.character(unlist(heads(x$Parent, 1)))

  mrna  <- gff[gff$type == "mRNA" & lengths(gff$Parent) > 0]
  tx2gene <- set_names(first_parent(mrna), mrna$ID)
  genes <- gff[gff$type == "gene" & gff$ID %in% tx2gene]
  names(genes) <- genes$ID

  ex <- gff[gff$type == "exon" & lengths(gff$Parent) > 0]
  ex <- ex[first_parent(ex) %in% names(tx2gene)]
  ex$gene_id <- tx2gene[first_parent(ex)]

  gene_body <- IRanges::reduce(genes, ignore.strand = TRUE)
  exons     <- IRanges::reduce(ex, ignore.strand = TRUE)
  introns   <- BiocGenerics::setdiff(gene_body, exons)
  upstream  <- trim(promoters(genes, upstream = promoter_bp, downstream = 0))
  promoters <- BiocGenerics::setdiff(IRanges::reduce(upstream, ignore.strand = TRUE), gene_body)
  intergenic <- gaps(IRanges::reduce(c(gene_body, promoters)))
  intergenic <- intergenic[strand(intergenic) == "*"]

  list(seqinfo = si, genes = genes,
       exons_by_gene = IRanges::reduce(split(granges(ex), ex$gene_id), ignore.strand = TRUE),
       features = list(exon = exons, intron = introns,
                       promoter = promoters, intergenic = intergenic))
}

annot <- map(set_names(seq_len(nrow(species)), sp_keys),
             ~ build_annotation(species$genome_fa[.x], species$gff[.x]))

map_dfr(annot, ~ tibble(
  genes       = length(.x$genes),
  genome_mb   = sum(as.numeric(seqlengths(.x$seqinfo))) / 1e6,
  exon_mb     = sum(width(.x$features$exon)) / 1e6,
  intron_mb   = sum(width(.x$features$intron)) / 1e6,
  promoter_mb = sum(width(.x$features$promoter)) / 1e6,
  intergenic_mb = sum(width(.x$features$intergenic)) / 1e6
), .id = "species")
```

# Genomic CpG o/e

CpG o/e = (CpG count × length) / (C count × G count). Germline methylation
deaminates methylated C to T over evolutionary time, so genes that are habitually
methylated are CpG-depleted (low o/e). In bivalves the gene distribution is often
bimodal, and that pattern is a prediction about the measured gbM below. It is
computed for both the gene body and the exons (CDS-dominated, more comparable across
annotations). CpG o/e is strand-symmetric, so strand is ignored.

```{r cpg-oe}
cpg_oe <- function(genome_fa, gr_list) {
  fa  <- FaFile(genome_fa)
  gr  <- unlist(gr_list, use.names = FALSE)
  ids <- rep(names(gr_list), lengths(gr_list))
  seqs <- getSeq(fa, gr)
  counts <- cbind(
    len = as.numeric(width(seqs)),
    CG  = oligonucleotideFrequency(seqs, 2)[, "CG"],
    letterFrequency(seqs, c("C", "G"))
  )
  agg <- rowsum(counts, ids)          # sum exon pieces per gene
  tibble(gene_id = rownames(agg),
         length  = agg[, "len"],
         cpg_oe  = agg[, "CG"] * agg[, "len"] / (agg[, "C"] * agg[, "G"]))
}

gene_oe <- map_dfr(set_names(seq_len(nrow(species)), sp_keys), function(i) {
  a <- annot[[i]]
  body <- split(granges(a$genes), names(a$genes))
  full_join(
    cpg_oe(species$genome_fa[i], body) %>% rename(gene_len = length, oe_gene = cpg_oe),
    cpg_oe(species$genome_fa[i], a$exons_by_gene) %>% rename(exon_len = length, oe_exon = cpg_oe),
    by = "gene_id")
}, .id = "species") %>%
  filter(is.finite(oe_exon))

write_tsv(gene_oe, file.path(out_dir, "gene_cpg_oe.tsv"))

gene_oe %>%
  pivot_longer(c(oe_gene, oe_exon), names_to = "region", values_to = "cpg_oe") %>%
  filter(cpg_oe < 2) %>%
  ggplot(aes(cpg_oe, fill = species)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~region) +
  labs(x = "CpG o/e", title = "Gene CpG o/e by species") + theme_bw()
```

# Per-CpG methylation

Each per-sample file is read once; per-oyster summaries are taken at
`sample_min_cov`, then counts are summed across oysters within a species to give a
pooled species-level methylome. Pooling is appropriate for the between-species
question: individual variation within a species is far smaller than the divergence
being compared here, and pooling recovers sites that no single low-depth oyster covers.

```{r read-cpg, eval = any(has_meth)}
read_cov <- function(f) {
  fread(f, header = FALSE, select = c(1, 2, 5, 6),
        col.names = c("chr", "start", "nC", "nT"))
}

per_sample <- list()
pooled <- list()
for (sp in sp_keys[has_meth]) {
  ss <- filter(samples, species == sp)
  d <- rbindlist(set_names(map(ss$cov_file, read_cov), ss$sample), idcol = "sample")
  d[, cov := nC + nT]

  per_sample[[sp]] <- d[cov >= sample_min_cov,
    .(cpgs = .N,
      global_meth_pct = 100 * sum(nC) / sum(cov),
      pct_meth_cpgs   = 100 * mean(100 * nC / cov >= meth_hi)),
    by = sample]

  pooled[[sp]] <- d[, .(nC = sum(nC), nT = sum(nT), n_samples = .N), by = .(chr, start)
                   ][nC + nT >= pooled_min_cov]
  rm(d); gc()
}

per_sample <- rbindlist(per_sample, idcol = "species") %>%
  left_join(select(samples, species, sample, group, assay), by = c("species", "sample"))
write_tsv(per_sample, file.path(out_dir, "per_sample_summary.tsv"))

ggplot(per_sample, aes(species, global_meth_pct, colour = group)) +
  geom_jitter(width = 0.15, size = 3) +
  labs(y = "Global CpG methylation (%)",
       subtitle = paste0("CpGs at >= ", sample_min_cov, "x per oyster")) +
  theme_bw()
```

```{r pooled-cpgs, eval = any(has_meth)}
cpg_gr <- map(pooled, function(p) {
  GRanges(p$chr, IRanges(p$start, width = 1),
          nC = p$nC, nT = p$nT, pct = 100 * p$nC / (p$nC + p$nT))
})

# Contig names in the CpG files must match the GFF/genome; catch a wrong assembly early.
iwalk(cpg_gr, function(gr, sp) {
  shared <- mean(as.character(seqnames(gr)) %in% seqlevels(annot[[sp]]$seqinfo))
  if (shared < 0.95)
    stop(sp, ": only ", round(100 * shared), "% of CpGs are on contigs in the genome ",
         "index - are the CpG calls from a different assembly?")
})

annotate_feature <- function(gr, features) {
  out <- rep(NA_character_, length(gr))
  for (f in names(features)) {
    hit <- is.na(out) & overlapsAny(gr, features[[f]], ignore.strand = TRUE)
    out[hit] <- f
  }
  factor(out, levels = names(features))
}

for (sp in names(cpg_gr)) {
  cpg_gr[[sp]]$feature <- annotate_feature(cpg_gr[[sp]], annot[[sp]]$features)
  cpg_gr[[sp]]$status  <- cut(cpg_gr[[sp]]$pct, c(-Inf, meth_lo, meth_hi, Inf),
                              right = FALSE,
                              labels = c("unmethylated", "sparse", "methylated"))
}

cpg_tbl <- imap_dfr(cpg_gr, ~ as_tibble(mcols(.x)[, c("pct", "feature", "status")]) %>%
                      mutate(species = .y))

cpg_tbl %>%
  group_by(species) %>%
  summarise(cpgs = n(), mean_pct = mean(pct),
            pct_methylated = 100 * mean(status == "methylated"),
            pct_unmethylated = 100 * mean(status == "unmethylated")) %>%
  write_tsv(file.path(out_dir, "methylome_summary.tsv")) %>%
  print()
```

## Distribution

```{r cpg-distribution, eval = any(has_meth)}
ggplot(cpg_tbl, aes(pct, after_stat(density), fill = species)) +
  geom_histogram(binwidth = 5, boundary = 0, position = "identity", alpha = 0.5) +
  labs(x = "% methylation per CpG (pooled)") + theme_bw()
```

## Feature partitioning

Where is methylation concentrated? Bivalves typically methylate gene bodies
(exons and introns) and leave promoters and intergenic DNA largely unmethylated.

```{r features-meth, eval = any(has_meth)}
feature_summary <- cpg_tbl %>%
  group_by(species, feature) %>%
  summarise(cpgs = n(), mean_pct = mean(pct),
            pct_methylated = 100 * mean(status == "methylated"), .groups = "drop") %>%
  group_by(species) %>%
  mutate(share_of_methylated = 100 * cpgs * pct_methylated / sum(cpgs * pct_methylated)) %>%
  ungroup()
write_tsv(feature_summary, file.path(out_dir, "feature_summary.tsv"))
feature_summary

feature_summary %>%
  pivot_longer(c(pct_methylated, share_of_methylated)) %>%
  ggplot(aes(feature, value, fill = species)) +
  geom_col(position = "dodge") +
  facet_wrap(~name, scales = "free_y",
             labeller = as_labeller(c(pct_methylated = "% of CpGs in feature methylated",
                                      share_of_methylated = "% of all methylated CpGs"))) +
  labs(x = NULL, y = NULL) + theme_bw()
```

# Gene-body methylation

Per gene: methylated / total calls summed over its covered CpGs (pooled across
oysters). Genes are split into methylated and unmethylated classes with a
two-group k-means on logit(gbM), separately per species, so the boundary adapts to
each assay rather than using one fixed cut-off.

```{r gene-body, eval = any(has_meth)}
gene_meth <- imap_dfr(cpg_gr, function(gr, sp) {
  ov <- findOverlaps(gr, annot[[sp]]$genes, ignore.strand = TRUE)
  tibble(gene_id = names(annot[[sp]]$genes)[subjectHits(ov)],
         nC = gr$nC[queryHits(ov)], nT = gr$nT[queryHits(ov)]) %>%
    group_by(gene_id) %>%
    summarise(n_cpg = n(), gbm_pct = 100 * sum(nC) / sum(nC + nT)) %>%
    filter(n_cpg >= min_gene_cpgs) %>%
    mutate(species = sp)
})

classify <- function(pct) {
  x  <- qlogis(pmin(pmax(pct / 100, 0.005), 0.995))
  km <- kmeans(x, centers = quantile(x, c(0.1, 0.9)))
  hi <- which.max(km$centers)
  factor(if_else(km$cluster == hi, "methylated", "unmethylated"),
         levels = c("unmethylated", "methylated"))
}

gene_meth <- gene_meth %>%
  group_by(species) %>%
  mutate(class = classify(gbm_pct)) %>%
  ungroup()
write_tsv(gene_meth, file.path(out_dir, "gene_body_methylation.tsv"))

gene_meth %>%
  group_by(species, class) %>%
  summarise(genes = n(), min_pct = min(gbm_pct), max_pct = max(gbm_pct),
            .groups = "drop") %>%
  group_by(species) %>% mutate(pct_genes = 100 * genes / sum(genes))

ggplot(gene_meth, aes(gbm_pct, fill = class)) +
  geom_histogram(binwidth = 2, boundary = 0) +
  facet_wrap(~species, ncol = 1, scales = "free_y",
             labeller = as_labeller(sp_labels)) +
  labs(x = "Gene-body methylation (%)", y = "Genes") + theme_bw()
```

## Measured gbM vs CpG o/e

Within each species, methylated genes should be CpG-depleted. A clear negative
relationship checks the data and justifies using CpG o/e as a proxy for the
species that lacks methylation calls.

```{r gbm-vs-oe, eval = any(has_meth)}
gbm_oe <- inner_join(gene_meth, gene_oe, by = c("species", "gene_id"))

gbm_oe %>%
  group_by(species) %>%
  summarise(genes = n(),
            rho_gene = cor(gbm_pct, oe_gene, method = "spearman"),
            rho_exon = cor(gbm_pct, oe_exon, method = "spearman"))

ggplot(filter(gbm_oe, oe_exon < 2), aes(oe_exon, gbm_pct)) +
  geom_bin2d(bins = 60) +
  geom_smooth(colour = "red", se = FALSE) +
  facet_wrap(~species, labeller = as_labeller(sp_labels)) +
  scale_fill_viridis_c(trans = "log10") +
  labs(x = "Exon CpG o/e", y = "Gene-body methylation (%)") + theme_bw()
```

# Orthologs

One-to-one orthologs are reciprocal best DIAMOND hits from `14.1`. Each comparison
uses all orthologs where both species have the measurement.

```{r orthologs}
rbh_file <- file.path(out_dir, "orthologs", paste0(sp_keys[1], "_", sp_keys[2], "_rbh.tsv"))
rbh <- read_tsv(rbh_file, show_col_types = FALSE)
gene_cols <- paste0(sp_keys, "_gene")
message(nrow(rbh), " one-to-one orthologs")

# One row per ortholog pair, with each species' metrics suffixed by species key.
wide <- function(tbl, cols) {
  per_sp <- map(sp_keys, function(sp) {
    tbl %>% filter(species == sp) %>%
      select(gene_id, all_of(cols)) %>%
      rename_with(~ paste0(.x, "_", sp), all_of(cols)) %>%
      rename(!!paste0(sp, "_gene") := gene_id)
  })
  rbh %>%
    select(all_of(gene_cols), pident) %>%
    left_join(per_sp[[1]], by = gene_cols[1]) %>%
    left_join(per_sp[[2]], by = gene_cols[2])
}

orth <- wide(gene_oe, c("oe_gene", "oe_exon"))
if (any(has_meth)) {
  orth <- orth %>%
    left_join(wide(gene_meth, c("n_cpg", "gbm_pct", "class")) %>%
                select(-pident), by = gene_cols)
}
write_tsv(orth, file.path(out_dir, "ortholog_methylation.tsv"))
```

## CpG o/e conservation

Available for every ortholog regardless of methylation data. A strong correlation
means the two lineages have methylated the same genes over the long term.

```{r orth-oe}
x_oe <- paste0("oe_exon_", sp_keys[1]); y_oe <- paste0("oe_exon_", sp_keys[2])
orth_oe <- orth %>% filter(.data[[x_oe]] < 2, .data[[y_oe]] < 2)
rho_oe <- cor(orth_oe[[x_oe]], orth_oe[[y_oe]], method = "spearman")

ggplot(orth_oe, aes(.data[[x_oe]], .data[[y_oe]])) +
  geom_bin2d(bins = 60) + geom_abline(linetype = 2) +
  scale_fill_viridis_c(trans = "log10") +
  labs(x = paste("Exon CpG o/e,", sp_labels[1]), y = paste("Exon CpG o/e,", sp_labels[2]),
       subtitle = sprintf("%d orthologs, Spearman rho = %.2f", nrow(orth_oe), rho_oe)) +
  theme_bw()
```

## Measured gbM conservation

Runs when both species have methylation calls.

```{r orth-gbm, eval = all(has_meth)}
x_m <- paste0("gbm_pct_", sp_keys[1]); y_m <- paste0("gbm_pct_", sp_keys[2])
x_c <- paste0("class_", sp_keys[1]);   y_c <- paste0("class_", sp_keys[2])
orth_m <- orth %>% filter(!is.na(.data[[x_m]]), !is.na(.data[[y_m]]))
rho_m <- cor(orth_m[[x_m]], orth_m[[y_m]], method = "spearman")

ggplot(orth_m, aes(.data[[x_m]], .data[[y_m]])) +
  geom_bin2d(bins = 50) + geom_abline(linetype = 2) +
  scale_fill_viridis_c(trans = "log10") +
  labs(x = paste("gbM (%),", sp_labels[1]), y = paste("gbM (%),", sp_labels[2]),
       subtitle = sprintf("%d orthologs, Spearman rho = %.2f", nrow(orth_m), rho_m)) +
  theme_bw()

# Class agreement: is a gene methylated in one species more likely methylated in the other?
class_tab <- table(orth_m[[x_c]], orth_m[[y_c]],
                   dnn = paste0(sp_keys, " class"))
class_tab
fisher.test(class_tab)
message(sprintf("Class agreement: %.1f%% of %d orthologs",
                100 * sum(diag(class_tab)) / sum(class_tab), sum(class_tab)))

# Genes whose methylation class differs between species: candidates for
# lineage-specific epigenetic change (check expression and annotation next).
orth_m %>%
  filter(.data[[x_c]] != .data[[y_c]]) %>%
  mutate(switch = paste0(sp_keys[1], "_", .data[[x_c]], "__", sp_keys[2], "_", .data[[y_c]]),
         delta_pct = .data[[y_m]] - .data[[x_m]]) %>%
  arrange(desc(abs(delta_pct))) %>%
  write_tsv(file.path(out_dir, "ortholog_class_switches.tsv")) %>%
  count(switch)
```

## Does conservation depend on sequence divergence?

Ortholog pairs binned by protein identity: if methylation tracks sequence
conservation, gbM agreement should rise with identity.

```{r orth-divergence, eval = all(has_meth)}
orth_m %>%
  mutate(identity_bin = cut(pident, c(0, 50, 70, 85, 100), include.lowest = TRUE)) %>%
  group_by(identity_bin) %>%
  summarise(orthologs = n(),
            rho = cor(.data[[x_m]], .data[[y_m]], method = "spearman"),
            class_agreement_pct = 100 * mean(.data[[x_c]] == .data[[y_c]]))
```

# Next steps

- **Function:** GO enrichment of conserved-methylated, conserved-unmethylated and
  class-switch orthologs against all tested orthologs, using the Swiss-Prot/GO join
  from `03` (Och gene IDs).
- **Expression:** join `ortholog_methylation.tsv` to the `09` limma tables on the Och
  gene ID, to ask whether genes that gained or lost gbM in *O. chilensis* differ in
  expression level or breadth.
- **Population layer:** both datasets have three populations. Once `13.4` is done,
  compare the within-species population DMGs (`DMG_*.tsv`) with an equivalent
  *O. lurida* analysis: are the same orthologs population-variable in both species?
- **Repeats/TEs:** add a RepeatMasker GFF per species to the feature set; TE
  methylation differs widely among bivalves and is not captured by the gene-centric
  features here.

```{r session}
sessionInfo()
```
