---
title: "15 - GN020540: a Rio-specific loss of methylation"
output:
  html_document:
    toc: true
    toc_float: true
date: "2026-10-10"
---

GN020540 is the strongest single population difference in `13.4`: the only gene
with a significant gene-body difference (Qui vs Rio), and covered by seven 1 kb DMRs
from the promoter to the last exon, all lower in Rio. This notebook looks at the
locus directly:

1. per-oyster methylation and read depth along the locus;
2. what the gene is (annotation) and whether it is expressed (ONT RNA-seq, `09`);
3. SNPs from `13.7` in the region;
4. the two haplotype assemblies (HapA, HapB) at the locus;
5. methylation at the other genomic copies of the repeat inside the intron, to tell
   a locus-specific (cis) change from a repeat-family-wide one.

**Upstream:** `13.2` (strand-merged CpG calls), `13.4`, `13.5` (Swiss-Prot hits),
`13.7` (SNPs). The haplotype assemblies are downloaded to `data/` if absent (as in
`12-compare-genomes.py`). RNA-seq BAMs and counts are read from gannet, where `bu.sh`
backs up `output/09-align-v1-clean/`.

**Samples:** the 9 independent oysters in `data/13-wgbs-samples.tsv` (duplicated
libraries excluded; see README Known issues).

**Where to run:** a compute node, inside the lab R container (~10 min, 16 cores for
minimap2). Command-line tools are called by absolute path from the srlab conda envs,
which run inside the container; paths are relative to `code/`.

**Network.** klone compute nodes have no internet access. The RNA-seq inputs (count
matrix, per-library read counts from the remote BAMs) and the haplotype assemblies
are fetched only if not already cached in `out_dir` / `data/`. Run
`code/15-GN020540-fetch.R` once from a login node (lab R container) to fill the cache
before knitting on a compute node.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 4.5)
```

```{r packages}
library(tidyverse)
```

```{r parameters}
out_dir  <- "../output/15-GN020540-locus"
meth_dir <- "../output/13-wgbs/methylation"
genome   <- "../data/13-wgbs-genome/merged_out.fasta"
gff_file <- "../data/13-wgbs-genome/GN.gene.gff3"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

gene_id  <- "GN020540"
chrom    <- "Chromosome_9B"
view     <- c(7180000, 7260000)   # wide window for depth
locus    <- c(7215000, 7234000)   # gene +/- ~6 kb
promoter_bp <- 2000

envs     <- "/mmfs1/gscratch/srlab/sr320/miniforge3/envs"
minimap2 <- file.path(envs, "myflow/bin/minimap2")
samtools <- file.path(envs, "wgbs-snp/bin/samtools")
bcftools <- file.path(envs, "wgbs-snp/bin/bcftools")
bedtools <- file.path(envs, "wgbs-snp/bin/bedtools")
threads  <- 16

gannet <- "https://gannet.fish.washington.edu/v1_web/owlshell/bu-github"
Sys.setenv(MINIMAP2 = minimap2, SAMTOOLS = samtools, BEDTOOLS = bedtools,
           OUT = out_dir, METH = meth_dir, GENOME = genome, CHROM = chrom,
           VLO = view[1], VHI = view[2], THREADS = threads)
```

# Gene model and annotation

```{r gene}
gff <- read_tsv(gff_file, comment = "#", col_names = FALSE, show_col_types = FALSE)
gene_rows <- gff %>% filter(str_detect(X9, paste0("(ID|Parent)=", gene_id, "(;|\\.|$)")))
gene  <- gene_rows %>% filter(X3 == "gene") %>% transmute(start = X4, end = X5, strand = X7)
exons <- gene_rows %>% filter(X3 == "exon") %>% transmute(start = X4, end = X5)
gene; exons

# Nearest other genes
gff %>% filter(X1 == chrom, X3 == "gene", X5 > gene$start - 50000, X4 < gene$end + 50000) %>%
  transmute(gene = str_extract(X9, "ID=[^;]+"), start = X4, end = X5)

annot <- read_tsv("../output/13-wgbs/13.5-gene-annotation/GN_gene_annotation.tsv",
                  show_col_types = FALSE)
annot %>% filter(gene_id == !!gene_id) %>%
  select(gene_id, entry_name, protein_name, pident, aln_len, qlen, slen, evalue)
```

The predicted protein (`qlen`) is compared with the full-length Swiss-Prot hit
(`slen`): a much shorter query means a partial gene model or a gene fragment.

# Methylation and depth per oyster

```{r samples}
samples <- read_tsv("../data/13-wgbs-samples.tsv", show_col_types = FALSE) %>%
  select(sample, population)
writeLines(samples$sample, file.path(out_dir, "samples.txt"))
```

Per-oyster CpG calls in the wide window, and each oyster's genome-wide mean depth
per covered CpG (to express local depth relative to the genome).

```{r extract-cpgs, engine='bash'}
while read -r s; do
  (
    f="${METH}/${s}.CpG_report.merged_CpG_evidence.cov.gz"
    zcat "$f" | awk -F'\t' -v OFS='\t' -v c="$CHROM" -v lo="$VLO" -v hi="$VHI" -v s="$s" \
      '$1 == c && $2 >= lo && $2 <= hi {print s, $2, $5, $6}' > "${OUT}/cpg_${s}.tsv"
    zcat "$f" | awk -F'\t' -v s="$s" '{n++; d += $5 + $6} END {print s "\t" d / n}' \
      > "${OUT}/depth_${s}.tsv"
  ) &
done < "${OUT}/samples.txt"
wait
cat "${OUT}"/cpg_*.tsv  > "${OUT}/cpg_window.tsv"
cat "${OUT}"/depth_*.tsv > "${OUT}/genome_depth.tsv"
wc -l "${OUT}/cpg_window.tsv"
```

```{r methylation-table}
cpg <- read_tsv(file.path(out_dir, "cpg_window.tsv"), col_names = c("sample", "pos", "nC", "nT"),
                show_col_types = FALSE) %>%
  left_join(samples, by = "sample") %>%
  mutate(cov = nC + nT)
gdepth <- read_tsv(file.path(out_dir, "genome_depth.tsv"), col_names = c("sample", "genome_depth"),
                   show_col_types = FALSE)

region_of <- function(pos) case_when(
  pos < gene$start - promoter_bp ~ "upstream flank",
  pos < gene$start ~ "promoter (2 kb)",
  pos <= gene$end ~ "gene body",
  TRUE ~ "downstream flank")

by_region <- cpg %>%
  filter(between(pos, locus[1], locus[2])) %>%
  mutate(region = factor(region_of(pos), levels = c("upstream flank", "promoter (2 kb)",
                                                    "gene body", "downstream flank"))) %>%
  group_by(population, sample, region) %>%
  summarise(cpgs = n(), meth_pct = 100 * sum(nC) / sum(cov), mean_depth = mean(cov),
            .groups = "drop") %>%
  left_join(gdepth, by = "sample") %>%
  mutate(rel_depth = mean_depth / genome_depth)
write_tsv(by_region, file.path(out_dir, "methylation_by_region.tsv"))

by_region %>% select(sample, region, meth_pct) %>%
  pivot_wider(names_from = region, values_from = meth_pct) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))
```

Per-CpG states in the gene body: are oysters near 0 / 100% (allelic or fixed) or
intermediate?

```{r cpg-states}
cpg %>%
  filter(between(pos, gene$start, gene$end), cov >= 3) %>%
  mutate(state = cut(100 * nC / cov, c(-1, 20, 80, 100), labels = c("<=20%", "20-80%", ">80%"))) %>%
  count(population, sample, state) %>%
  group_by(sample) %>% mutate(frac = round(n / sum(n), 2)) %>% select(-n) %>%
  pivot_wider(names_from = state, values_from = frac, values_fill = 0)
```

```{r profile-plot}
bins <- cpg %>%
  filter(between(pos, locus[1], locus[2])) %>%
  mutate(bin = floor(pos / 500) * 500 + 250) %>%
  group_by(population, sample, bin) %>%
  summarise(meth = 100 * sum(nC) / sum(cov), cov = sum(cov), .groups = "drop") %>%
  filter(cov >= 5)
dmr <- read_tsv("../output/13-wgbs/13.4-methylation/DMR_Qui_vs_Rio.tsv", show_col_types = FALSE) %>%
  filter(seqnames == chrom, end >= locus[1], start <= locus[2])

p <- ggplot(bins, aes(bin / 1e6, meth, colour = population, group = sample)) +
  geom_rect(data = dmr, inherit.aes = FALSE,
            aes(xmin = start / 1e6, xmax = end / 1e6, ymin = -Inf, ymax = Inf),
            fill = "grey90") +
  geom_rect(data = exons, inherit.aes = FALSE,
            aes(xmin = start / 1e6, xmax = end / 1e6, ymin = -8, ymax = -3), fill = "black") +
  annotate("segment", x = gene$start / 1e6, xend = gene$end / 1e6, y = -5.5, yend = -5.5) +
  geom_line(alpha = 0.8) + geom_point(size = 0.8) +
  scale_colour_manual(values = c(Pum = "grey40", Qui = "#1b9e77", Rio = "#d95f02")) +
  labs(x = paste(chrom, "(Mb)"), y = "CpG methylation (%), 500 bp bins",
       title = paste(gene_id, "methylation per oyster"),
       subtitle = "Grey = 13.4 DMR tiles; black = exons") +
  theme_bw()
p
ggsave(file.path(out_dir, paste0(gene_id, "_methylation.png")), p, width = 9, height = 4.5, dpi = 150)
```

Read depth relative to each oyster's genome-wide mean, in 4 kb bins over the wide
window. Depth well above 1 in every oyster points to sequence present more than once
in the genome but once in the assembly; a Rio-only drop would point to a deletion.

```{r depth}
depth_bins <- cpg %>%
  mutate(bin_kb = floor(pos / 4000) * 4) %>%
  group_by(sample, bin_kb) %>%
  summarise(depth = mean(cov), .groups = "drop") %>%
  left_join(gdepth, by = "sample") %>%
  mutate(rel = round(depth / genome_depth, 2)) %>%
  select(sample, bin_kb, rel) %>%
  pivot_wider(names_from = bin_kb, values_from = rel)
write_tsv(depth_bins, file.path(out_dir, "relative_depth_4kb.tsv"))
depth_bins
```

# SNPs in the region (13.7)

```{r snps}
vcf_q20 <- "../output/13-wgbs/13.7-snps/wgbs_snps.q20.vcf.gz"
vcf_all <- "../output/13-wgbs/13.7-snps/wgbs_snps.all.vcf.gz"
region  <- sprintf("%s:%d-%d", chrom, locus[1], locus[2])
q20 <- system2(bcftools, c("view", "-H", "-r", region, vcf_q20), stdout = TRUE)
all_var <- system2(bcftools, c("query", "-r", region, "-i", "'ALT!=\".\"'", "-f", "'%QUAL\\n'", vcf_all),
                   stdout = TRUE)
tibble(region = region, snps_qual20 = length(q20),
       candidate_variant_sites = length(all_var),
       max_qual = suppressWarnings(max(as.numeric(all_var))))
```

# Expression (ONT RNA-seq, 09)

StringTie counts from `09` (current pipeline), and direct read counts at the gene
from each library's BAM, read remotely through its index. The RNA-seq oysters are
different individuals from the WGBS oysters.

```{r expression-counts}
counts_file <- file.path(out_dir, "09_gene_count_matrix.csv")
if (!file.exists(counts_file))
  download.file(file.path(gannet, "project-ostrea-chil/output/09-align-v1-clean/stringtie_quant/gene_count_matrix.csv"),
                counts_file, quiet = TRUE)
counts <- read_csv(counts_file, show_col_types = FALSE)
libsize <- colSums(select(counts, -gene_id))
counts %>% filter(str_detect(gene_id, gene_id)) %>%
  pivot_longer(-gene_id, names_to = "library", values_to = "count") %>%
  mutate(cpm = round(1e6 * count / libsize[library], 1))

# How common is "zero in half the libraries" among expressed genes?
m <- as.matrix(select(counts, -gene_id))
expressed <- rowSums(m) >= 100
c(expressed_genes = sum(expressed), zero_in_6_plus = sum(rowSums(m[expressed, ] == 0) >= 6))
```

```{r expression-bams}
libs <- setdiff(names(counts), "gene_id")
bam_url <- function(l) file.path(gannet, "project-ostrea-chil/output/09-align-v1-clean",
                                 paste0(l, "_combined_clean_porechop.sorted.bam"))
gene_region <- sprintf("%s:%d-%d", chrom, gene$start, gene$end)
reads_file <- file.path(out_dir, "rnaseq_reads_at_gene.tsv")
if (file.exists(reads_file)) {
  reads <- read_tsv(reads_file, show_col_types = FALSE)
} else {
old <- setwd(out_dir)   # samtools caches the remote .bai files in the working directory
reads <- map_dfr(libs, function(l) {
  idx <- system2(samtools, c("idxstats", bam_url(l)), stdout = TRUE)
  mapped <- sum(as.numeric(map_chr(strsplit(idx, "\t"), 3)))
  n <- as.integer(system2(samtools, c("view", "-c", "-F", "0x904", "-q", "20",
                                      bam_url(l), gene_region), stdout = TRUE))
  tibble(library = l, population = substr(l, 1, 3), mapped = mapped, gene_reads = n,
         per_million = round(1e6 * n / mapped, 1))
})
setwd(old)
write_tsv(reads, reads_file)
}
reads
reads %>% group_by(population) %>% summarise(mean_per_million = mean(per_million))
```

# Haplotype assemblies

The merged reference's `Chromosome_9B` is HapB's `Chromosome_9B`. The 50 kb around
the gene, and separately the high-depth intron segment, are aligned to HapA, HapB and
the merged assembly.

```{r haplotypes-download}
hap_urls <- c(HapA = file.path(gannet, "OCEAN/docs/jbrowse/data/HapA/Och_HapA_assembly.fa"),
              HapB = file.path(gannet, "OCEAN/docs/jbrowse/data/HapB/Och_HapB_assembly.fa"))
hap_files <- c(HapA = "../data/Och_HapA_assembly.fa", HapB = "../data/Och_HapB_assembly.fa")
for (h in names(hap_files)) {
  if (!file.exists(hap_files[h])) {
    download.file(hap_urls[h], paste0(hap_files[h], ".part"), mode = "wb", quiet = TRUE)
    file.rename(paste0(hap_files[h], ".part"), hap_files[h])
  }
}
file.size(hap_files) / 1e6
```

```{r haplotypes-align, engine='bash'}
"$SAMTOOLS" faidx "$GENOME" "${CHROM}:7200000-7250000" > "${OUT}/locus50k.fa"
"$SAMTOOLS" faidx "$GENOME" "${CHROM}:7222000-7227000" > "${OUT}/intron_segment.fa"
for t in HapA:../data/Och_HapA_assembly.fa HapB:../data/Och_HapB_assembly.fa merged:"$GENOME"; do
  n=${t%%:*}; f=${t#*:}
  [[ -s "${OUT}/locus_vs_${n}.paf" ]] || \
    "$MINIMAP2" -t "$THREADS" -x asm20 -c --secondary=yes -N 200 -p 0.1 "$f" "${OUT}/locus50k.fa" \
      > "${OUT}/locus_vs_${n}.paf" 2> /dev/null
  [[ -s "${OUT}/intron_vs_${n}.paf" ]] || \
    "$MINIMAP2" -t "$THREADS" -x asm20 -c --secondary=yes -N 500 -p 0.05 "$f" "${OUT}/intron_segment.fa" \
      > "${OUT}/intron_vs_${n}.paf" 2> /dev/null
done
wc -l "${OUT}"/*.paf
```

```{r haplotypes-summary}
read_paf <- function(f) {
  read_tsv(f, col_names = FALSE, show_col_types = FALSE, col_select = 1:13) %>%
    transmute(qstart = X3, qend = X4, strand = X5, target = X6, tstart = X8, tend = X9,
              identity = round(X10 / X11, 3), aln_len = X11, mapq = X12,
              type = str_remove(X13, "tp:A:"))
}
offset <- 7200000   # query coordinates are relative to this; gene at q 21,497-27,996

# Primary placement of the 50 kb locus in each haplotype
map_dfr(c("HapA", "HapB"), ~ read_paf(file.path(out_dir, paste0("locus_vs_", .x, ".paf"))) %>%
          filter(type == "P") %>% mutate(assembly = .x)) %>%
  mutate(q = paste0(qstart + offset, "-", qend + offset)) %>%
  select(assembly, q, strand, target, tstart, tend, identity, aln_len)

# Self-similar blocks upstream of the gene in HapB (tandem repeat array)
read_paf(file.path(out_dir, "locus_vs_HapB.paf")) %>%
  filter(type == "S", target == chrom, aln_len >= 5000) %>%
  arrange(qstart) %>% select(qstart, qend, tstart, tend, identity, aln_len)

# Copies of the intron segment (>= 2 kb aligned)
map_dfr(c("HapA", "HapB", "merged"), ~ read_paf(file.path(out_dir, paste0("intron_vs_", .x, ".paf"))) %>%
          filter(aln_len >= 2000) %>%
          summarise(copies = n(), copies_95pct = sum(identity >= 0.95)) %>% mutate(assembly = .x))
```

# Other copies of the intronic repeat

If the Rio loss were a property of the repeat family (e.g. genome-wide repeat
methylation differing by population), the other copies would also be lower in Rio.

```{r repeat-copies, engine='bash'}
awk -F'\t' -v OFS='\t' -v c="$CHROM" \
  '$11 >= 2000 && !($6 == c && $8 < 7230000 && $9 > 7215000) {print $6, $8, $9}' \
  "${OUT}/intron_vs_merged.paf" | sort -k1,1 -k2,2n | "$BEDTOOLS" merge -i - > "${OUT}/repeat_copies.bed"
cut -f1,2 "${GENOME}.fai" | sort -k1,1 > "${OUT}/genome.sorted.sizes"
while read -r s; do
  (
    zcat "${METH}/${s}.CpG_report.merged_CpG_evidence.cov.gz" \
      | awk -F'\t' -v OFS='\t' '{print $1, $2 - 1, $2, $5, $6}' \
      | "$BEDTOOLS" intersect -sorted -g "${OUT}/genome.sorted.sizes" -a - -b "${OUT}/repeat_copies.bed" -wa -wb \
      | awk -F'\t' -v OFS='\t' -v s="$s" '{print s, $6 ":" $7, $4, $5}' > "${OUT}/repeat_${s}.tsv"
  ) &
done < "${OUT}/samples.txt"
wait
wc -l < "${OUT}/repeat_copies.bed"
```

```{r repeat-copies-summary}
rep <- map_dfr(samples$sample, ~ read_tsv(file.path(out_dir, paste0("repeat_", .x, ".tsv")),
                                          col_names = c("sample", "copy", "nC", "nT"),
                                          show_col_types = FALSE)) %>%
  left_join(samples, by = "sample")

per_copy <- rep %>% group_by(population, sample, copy) %>%
  summarise(calls = sum(nC + nT), meth = 100 * sum(nC) / calls, .groups = "drop") %>%
  filter(calls >= 20)

per_copy %>% group_by(population, sample) %>%
  summarise(copies = n(), mean_per_copy = round(mean(meth), 1),
            copies_below_10pct = sum(meth < 10), .groups = "drop")

# The GN020540 copy itself, for comparison (intron segment only)
cpg %>% filter(between(pos, 7222000, 7227000)) %>%
  group_by(population, sample) %>%
  summarise(meth_pct = round(100 * sum(nC) / sum(cov), 1), .groups = "drop")
```

# Summary

- Rio oysters are near-unmethylated across GN020540 (promoter to last exon); Quilhua
  and Pum1 are partly methylated, with no overlap between the groups.
- The gene model is a fragment (`qlen` vs `slen` above), and the intron holds a
  ~4 kb interspersed repeat with hundreds of genomic copies, which explains the
  doubled read depth and the intermediate methylation there.
- No confident SNP lies in the region; ONT expression is too sparse to compare.
- HapA carries an identical copy of the gene; the haplotypes differ by a tandem
  array just upstream, which short bisulfite reads cannot genotype.
- Other copies of the intronic repeat are methylated the same in every population,
  so the Rio loss is specific to this locus.

```{r session}
sessionInfo()
```
